\chapter{Methodology}
\label{chap:method}

% See context/outline.md Ch 3 for paragraph-level plan.


\section{Defining the Task}
\label{sec:defining-task}

RP began as a bachelor task offered by Admmit, a Norwegian software firm building products for the transport sector. The task description, distributed through NTNU as project OPG29, set out a brief: investigate how Norwegian traffic coordinators plan daily transport assignments, and build a working prototype of a planning platform that could assist them. The team accepted the task and committed to delivering both the artefact and the thesis that surrounds it. From that decision onwards, two roles ran in parallel for the two-student team: building RP as a functioning platform, and conducting the research that would explain and justify its design choices. Both students contributed to both workstreams over the course of the project.

The brief defined the topic but not the empirical foundation, and that part of the work was arranged by the team rather than by Admmit. Seven traffic coordinators and transport managers were contacted directly by phone over a single day in March 2026. All seven worked at companies that were already Admmit customers, but Admmit did not broker the introductions. The team found the coordinators through Admmit's customer roster, called them on its own initiative, and conducted brief semi-structured interviews lasting roughly fifteen minutes each. The interviewed companies range from a small operator with eight to twenty drivers to a multi-department fleet of around forty-five vehicles, and they use a mix of Timpex, Opter, internal tools, and fully manual planning.

Before any coordinator was on the phone, however, two design choices had already been locked. Admmit specified Human-in-the-Loop (HITL) operation as a non-negotiable property of the artefact from the project's first week: the algorithm proposes a plan; the coordinator inspects, modifies, accepts, or rejects each assignment. Admmit also specified a multi-tenant architecture in which a single deployed instance serves multiple customer companies through configuration. Both decisions originate in Admmit's mandate, not in the interviews. The interviews validated HITL by surfacing the coordinator's insistence on retaining authority over assignments, and they validated the cross-company variety that motivates configurable behaviour, but they did not introduce either property. The origin map presented later in this chapter (\Cref{sec:iterations}) labels every major design decision by origin and by interview-validation status, so the reader can tell apart Admmit-mandated choices, interview-driven choices, and designer / domain-knowledge choices.

This origin frames the rest of the methodology. Every design choice traced through the chapters that follow comes from one of three sources: the Admmit mandate that fixed HITL and multi-tenancy at the outset, the dialogue with the seven coordinators that informed the planning interface and the assignment-criteria taxonomy, or designer and domain-knowledge decisions that filled in what neither the mandate nor the interviews specified. HITL turns the \textbf{Trust/control} anchor into four concrete coordinator actions on every algorithm-generated assignment. The multi-tenant architecture supports \textbf{Adaptability} by isolating per-customer configuration in a single deployed instance. The interview-driven analysis identifies the \textbf{Efficiency} anchor as the resource-utilization visibility gap that the artefact's planning interface and benchmark are designed to address. The remaining sections of this chapter describe the research paradigm under which the artefact was built (\Cref{sec:dsr-dsrm}) and how the interview and analysis work was conducted (\Cref{sec:data-collection,sec:data-analysis}). The eight named iterations that produced the system are documented in \Cref{sec:iterations}. The framework under which it was evaluated, together with the validity bounds that follow, is set out in \Cref{sec:evaluation-framework,sec:validity}.


\section{Method Theory: DSR + DSRM Applied}
\label{sec:dsr-dsrm}

\Cref{sec:defining-task} mapped where each design choice originated; this section names the research paradigm under which those choices were made. The DSR foundation, together with the positivist and interpretive alternatives it is compared against, is set out in \Cref{sec:dsr}, and that comparison is not repeated here. A purely positivist design would require deployment-scale measurement of overtime and idle-time effects in Norwegian transport companies, which the present project's stage does not support. A purely interpretive design would yield rich description of coordinator practice without producing the runnable platform the bachelor task required. DSR is the paradigm the project requires (see \Cref{sec:dsr}), and RP is precisely such an artefact.

The methodology used in this thesis is \textcite{peffers2007dsrm}'s Design Science Research Methodology (DSRM), which structures DSR as six nominally sequential activities: problem identification and motivation; objectives for a solution; design and development; demonstration; evaluation; and communication \parencite[p.~46]{peffers2007dsrm}. The activities are nominally sequential rather than strictly so; \textcite{peffers2007dsrm} note explicitly that researchers are not expected to proceed in order from activity~1 through activity~6 \parencite[p.~56]{peffers2007dsrm}. The six activities below describe what was done in this project at each step.

\begin{description}
  \item[Problem identification and motivation.] The seven coordinator interviews surfaced a resource-utilization visibility gap: overtime is handled reactively, idle time between assignments is invisible, and load balancing depends on coordinator memory rather than any structured overview. Admmit's bachelor task framed the same gap from the owner side, asking for an algorithm-assisted planning platform that could make utilisation legible and adjustable. \Cref{sec:interview-findings} describes the empirical gap in detail; \Cref{sec:efficiency} discusses it under the Efficiency anchor.
  \item[Objectives for a solution.] The objectives are stated under the three locked anchor concepts. \textbf{Efficiency} is realised by the artefact reducing overtime, reducing idle time between assignments, and reducing uneven load between drivers. \textbf{Trust/control} is realised as the coordinator's authority to inspect, modify, accept, or reject any algorithm-generated assignment. \textbf{Adaptability} is delivered by configurable constraint weights and per-company assignment rules within a single deployed instance that can serve companies with different operational rules.
  \item[Design and development.] The artefact is an algorithm-assisted planning platform built on Next.js (a React-based web framework), tRPC (a typed remote-procedure-call layer between the client and the server), and PostgreSQL (a relational database). Its solver layer integrates three engines invoked as subprocesses: a greedy constructive heuristic, the OR-Tools CP-SAT solver (Google's open-source constraint-programming engine), and Timefold (an open-source metaheuristic constraint-solver platform), hereafter called the three engines and detailed in \Cref{sec:iter-multiengine}. A drag-and-drop timeline lets the coordinator inspect, modify, accept, or reject each assignment. The platform was built across eight named iterations, each labelled by origin (Admmit-mandated, interview-driven, or designer / domain-knowledge), described in \Cref{sec:iterations}.
  \item[Demonstration.] Demonstration takes the form of a multi-engine benchmark on synthetic datasets sized to cover the realistic fleet ranges of the seven interviewed companies, from small fleets of around ten drivers up to multi-department operations of around forty-five vehicles. The benchmark shows how the three engines behave when given the same problem instances. The dataset design and benchmark protocol are set out in \Cref{sec:evaluation-framework}.
  \item[Evaluation.] What the artefact is validated against is how solver approaches compare under realistic constraint combinations, not whether the visibility gap is real or whether the HITL stance is necessary. The benchmark validates solver performance independently of those whether-questions; the requirements traceability matrix records, for each functional and non-functional requirement, whether it was implemented and how its behaviour was checked. What is not validated in this thesis is documented in \Cref{sec:limitations}: the artefact is not deployed in production, real-world utilisation gains are not measured, and the override flow is not user-tested with coordinators. The validation-versus-evaluation distinction that bounds these claims is set out under DSR theory in \Cref{sec:dsr} and applied to the benchmark in \Cref{sec:evaluation-framework}.
  \item[Communication.] The research is communicated through two channels. The first is this thesis, addressed to the academic audience at NTNU and presenting RP as an instantiation in DSR's sense alongside the design knowledge embedded in its multi-engine and HITL choices. The second is a written recommendation to Admmit covering deployment readiness, including the requirements not yet validated through user testing (\Cref{sec:limitations}), and the per-company configurability through which Adaptability is delivered.
\end{description}

The mapping above places the project on Peffers' nominal sequence, but the actual ordering reflected the project's entry point rather than the sequence. \textcite{peffers2007dsrm} note four entry points into the DSRM cycle, and explicitly identify objective-centered initiation, in which an industry or research need fixes key objectives before activity~1 is complete, as a legitimate starting point \parencite[p.~56]{peffers2007dsrm}. RP fits that pattern: Admmit's mandate fixed HITL and the multi-tenant architecture (the artefact elements that realise Trust/control and Adaptability) before activity~1 was complete, and the eight iterations of activity~3 ran in parallel with later refinements of the objectives.


\section{Data Collection}
\label{sec:data-collection}

The Problem-Identification activity in \Cref{sec:dsr-dsrm} drew its empirical input from a single instrument: brief semi-structured interviews with traffic coordinators and transport managers across seven Norwegian transport companies. Semi-structured was chosen because the activity needed both qualities at once: data open enough to surface unanticipated themes, and structured enough to compare the same five guide topics across the seven calls. The visibility gap and the operator-vs-owner asymmetry both emerged this way rather than from any prepared question. A fully structured questionnaire would have closed off unanticipated themes; a fully open conversation would have lost cross-case comparability. The framing for the analytic step that turned these interviews into themes is set out in \Cref{sec:data-analysis}.

With the form of the instrument fixed, the next decision was who to consult. The seven coordinators were selected purposively from Admmit's customer roster rather than recruited through a public call, because the project required coordinators in the planning role at companies where algorithm-assisted planning is plausibly relevant, and Admmit's customer base supplied that population directly. The companies span small operators with around eight to twenty drivers through to a multi-department fleet of around forty-five vehicles, range geographically from Bergen and Mo i Rana to Lakselv and Tana, and cover the systems used in the field: Timpex, Opter, internal or custom tools, and fully manual planning. The self-selection that follows from the Admmit-customer recruitment route is a known limitation on the empirical foundation and is named as L2 in \Cref{sec:limitations}.

Within those seven calls, the same five guide topics anchored every conversation: current tools and workflow, assignment criteria, sick-leave handling, attitudes to automation, and adoption criteria. Each topic opened with a broad question and was followed up with prompts only where the coordinator's first answer left an obvious gap; this kept the calls comparable on topic coverage while preserving the open-question structure the form requires. Together, these five topics and the broad-question / follow-up-only-on-gaps convention served as the topic guide; no separate scripted questionnaire was used beyond the structure described here.

The seven calls were conducted by phone over a single day in March 2026, in Norwegian, with each call lasting approximately fifteen minutes for a cumulative interview time of roughly 105 minutes. Phone was the practical choice given the geographic spread of the coordinators (Bergen, Mo i Rana, Lakselv, Tana) and the brief format the consultation could realistically ask of working coordinators recruited cold. Norwegian was chosen as the working language because all seven coordinators were native speakers, and the topic vocabulary (TMS terminology, regulatory references, sick-leave handling) is naturally expressed in Norwegian. The fifteen-minute target balanced cross-topic coverage against the realistic attention budget of a working day. The single-day window kept the interviewer's framing consistent across calls and avoided the drift that comes from conducting interviews weeks apart, but it also constrained how deeply any one topic could be probed within a single call; the brief duration is named alongside the sample size as L1 in \Cref{sec:limitations}.

Recordings were transcribed automatically and the segments used for analysis were manually corrected against the audio. Automated transcription was chosen because the cumulative audio of around 105 minutes made fully manual transcription disproportionate to the analytic depth the consultations could support.

The interviews were conducted as an industry consultation, which has ethical consequences worth stating directly. The project ran under Admmit's bachelor task brief and NTNU institutional supervision rather than as formal research, and on that basis no Sikt (Norway's national agency for shared services in research and education) registration was sought. The conversations covered work practices, system use, and assignment criteria only; no special-category personal data (in the GDPR sense, covering health, ethnicity, political views, and similar) was collected. Consent was implicit: participants agreed to take the call and answer questions. The consultation framing makes this acceptable given the absence of sensitive data and the professional context, but no formal consent form was administered and recording proceeded without an explicit pre-recording consent step. No formal anonymisation protocol was applied for the same reason: the seven coordinators are professionals speaking about their working practices rather than research participants whose personal data is being processed. Only Timpex and Opter, the two commercial vendors named in the consultation pool, are named in the thesis, while the seven interviewed companies are referenced generically by size, sector, and system maturity. Recordings and transcriptions are stored locally on the project author's device, accessible only to the author and supervisor, and will be deleted after submission and grading. The researcher--developer overlap and the informal study procedures that follow from this consultation framing are named together as L3 in \Cref{sec:limitations}.


\section{Data Analysis}
\label{sec:data-analysis}

The seven interviews described in \Cref{sec:data-collection} produced roughly 105 minutes of audio across the five guide topics that anchored each call (current tools and workflow, assignment criteria, sick-leave handling, attitudes to automation, and adoption criteria). The analytic instrument applied to that material was thematic analysis as set out by \textcite{braun2006thematic}, chosen for its theoretical flexibility within the DSR frame \parencite[p.~81]{braun2006thematic} and its accessibility to a small research team conducting analysis as one component of a larger DSR Problem-Identification activity \parencite[p.~77]{braun2006thematic}. The analysis was conducted in the six phases Braun and Clarke specify: familiarisation, initial coding, theme search, theme review, theme definition, and report production \parencite[p.~87]{braun2006thematic}.

Three positioning choices situate the analysis within the framework \textcite{braun2006thematic} require analysts to make explicit \parencite[pp.~81--82]{braun2006thematic}: it was \emph{inductive}, in that no codebook was drawn from a prior hypothesis and the visibility gap and operator-vs-owner asymmetry were named only after coding; \emph{semantic}, in that codes worked at the level of coordinators' explicit statements about tools, criteria, sick-leave, automation, and adoption rather than at a latent discursive level; and \emph{essentialist/realist} (treating accounts as reports of working practice rather than as discursive constructions), in that coordinator accounts were treated as reports of real working practice usable as input to DSR Problem Identification (\Cref{sec:dsr-dsrm}). The analysis identified themes across the seven transcripts that elaborated and cut across the five guide topics, among them the resource-utilization visibility gap, the operator-vs-owner asymmetry, and automation scepticism with an override preference; following \textcite[Table~2, criterion~15]{braun2006thematic}, those themes were named through the coding process rather than treated as having emerged unaided.

Braun and Clarke's own examples are drawn from psychology and gender studies. The application here is to brief semi-structured consultations with traffic coordinators inside a DSR Problem-Identification activity, not to standalone qualitative research. The six-phase guide transfers across that domain gap by treating each transcript as a data item and the seven transcripts together as the data set.

The output of the analysis fed three later sections. The named themes are reported in \Cref{sec:interview-findings}. Each theme was then translated into one or more functional requirements, prioritised under MoSCoW (Must, Should, Could, Won't); MoSCoW was chosen because it makes the boundary between in-scope and out-of-scope requirements explicit and negotiable with stakeholders. The requirements are recorded in the functional-requirements register (IDs FK-01 through FK-42) and reported in \Cref{sec:requirements}: the visibility-gap theme produced the timeline view and conflict-detection requirements at Must; the assignment-criteria theme produced the hard- and soft-constraint requirements at Must and Should respectively; the automation-attitudes theme produced the override and weight-configuration requirements at Must and Should. A fit/gap analysis (a side-by-side comparison of what existing systems cover and what they do not) was then run against Timpex, Opter, and the internal or manual tools named in the consultation pool, producing five fit items where existing systems already meet the need and fourteen gap items where they do not, reported in \Cref{sec:fitgap}.

The analysis carries the limitations of its inputs and its frame, named as L1 and L2 in \Cref{sec:data-collection} and catalogued in full in \Cref{sec:limitations} rather than re-explained here. The themes are accordingly bounded to the seven interviewed companies and are not a basis for generalising to Norwegian transport more broadly.


\section{Iterative Development Process}
\label{sec:iterations}

The artefact described in \Cref{sec:dsr-dsrm} was built across eight named iterations, told here as a sequence of decisions rather than as a sprint diary. Each iteration follows the same pattern: what was tried, why it was tried, what happened (positive results and limitations), what was learned, and what motivated the next step. The opening iteration establishes the data substrate (\Cref{sec:iter-schema}); iterations two through four progress from a single greedy heuristic to a multi-engine solver layer (\Cref{sec:iter-greedy,sec:iter-cpsat,sec:iter-multiengine}); iterations five through seven build the HITL interface that lets the coordinator inspect, modify, accept, or reject every algorithm-generated assignment (\Cref{sec:iter-hitl,sec:iter-deviations,sec:iter-tradeoff}); and the eighth iteration exposes per-company configurability of the soft-constraint weight space (\Cref{sec:iter-weights}). Each iteration carries an inline origin label per the origin map introduced in \Cref{sec:defining-task}: Admmit-mandate, interview-validated, designer-technical, or a mixture. Iteration boundaries are reconstructed retrospectively from project artefacts and decision points; precise dates and ordering are approximate. The narrative below describes how the artefact was built; the framework under which it was tested follows in \Cref{sec:evaluation-framework}.

\subsection{Schema and Timeline Scaffolding}
\label{sec:iter-schema}

\begin{description}
  \item[Origin.] Admmit-mandate (multi-tenant architecture) combined with designer-technical exploration of the data model and a read-only timeline UI (user interface).
  \item[Tried.] A core entity model covering employees, vehicles, assignments, certifications, weekly work schedules, and time-off; multi-tenant data isolation enforced at the data layer with every query scoped by \texttt{companyId}; and a read-only Gantt-style timeline that rendered existing assignments without any optimisation step.
  \item[Why.] Later solver iterations need a structured substrate to operate over, and the multi-tenant boundary must be enforced at the data layer because Admmit's customer structure means each transport company would consume the artefact independently; bolting tenancy on later would risk cross-tenant leakage on every new query path.
  \item[What happened.] The substrate was adequate for the simplest assignment heuristic but required revision once constraint-solver formalisation surfaced gaps in the data model. Explicit time-off windows, certificate-expiry as a queryable date, and per-day work-schedule entries became necessary only after the heuristic exposed what the solver would need to reason about.
  \item[Learned.] Schema-first discipline reduces rework but cannot fully anticipate solver requirements. The substrate is itself an iteration, not a one-shot setup.
  \item[Next.] With the substrate stable enough to support a real assignment step, the next iteration attempted the simplest possible assignment heuristic.
\end{description}

\subsection{Single-engine Baseline (greedy)}
\label{sec:iter-greedy}

\begin{description}
  \item[Origin.] Designer-technical exploration.
  \item[Tried.] A priority-sorted greedy first-fit assignment heuristic. Assignments were sorted by priority, walked through in order, and each was given the first feasible driver and vehicle combination found.
  \item[Why.] An instant baseline that produces a feasible plan provides a reference point for measuring richer-solver gains and gives the coordinator immediate UI feedback while drag-and-drop edits happen.
  \item[What happened.] The heuristic worked on small instances of fewer than thirty assignments. On larger instances, priority-sorted greedy committed early to locally optimal but globally suboptimal choices, blocking later high-priority assignments and producing avoidable idle time and overtime.
  \item[Learned.] The greedy heuristic's quality limit is structural rather than tunable: reordering the priority comparator cannot break out of it because the heuristic lacks both constraint propagation and a global view of the day.
  \item[Next.] Introduce a complete solver that can reason about all assignments simultaneously.
\end{description}

\subsection{Constraint Generalisation (CP-SAT)}
\label{sec:iter-cpsat}

\begin{description}
  \item[Origin.] Designer-technical exploration.
  \item[Tried.] The Google OR-Tools CP-SAT solver with explicit hard constraints (competencies, availability, no double-booking, vehicle type) and weighted soft constraints (workload balance, vehicle preference, transitions, priority, employee preferences) combined under an objective function maximising scheduled assignments minus weighted soft-constraint violations.
  \item[Why.] Constraint programming is well-suited to assignment problems with combinations of hard and soft constraints, and CP-SAT specifically returns near-optimal solutions within a configurable time budget, a property that matters for an HITL interface where the coordinator is waiting on the solver.
  \item[What happened.] The solver returned near-optimal solutions on small to medium instances within a thirty-to-sixty-second time limit. Quality degraded on larger instances when the time limit bound: the solver returned the best feasible solution found rather than the optimum. Modelling complexity was higher than for the heuristic, and several iterations were needed to get the hard-versus-soft mapping correct.
  \item[Learned.] CP-SAT covers the medium-instance gap that the greedy heuristic could not, but introduces a scaling boundary that motivates a third engine for larger instances.
  \item[Next.] Add a third engine for large instances and a mechanism for comparing the three under matched conditions.
\end{description}

\subsection{Multi-engine Selection and Benchmarking}
\label{sec:iter-multiengine}

\begin{description}
  \item[Origin.] Designer-technical exploration. The multi-engine architecture lets the benchmark test how solvers compare without having to test whether the visibility gap is real or whether HITL is necessary; that separation is the framing carried through to \Cref{sec:evaluation-framework}.
  \item[Tried.] A pluggable solver registry that instantiates each of the three solver families introduced in \Cref{sec:scheduling} as one concrete engine: constructive heuristics as a greedy baseline, constraint programming as OR-Tools CP-SAT, and metaheuristics as Timefold (tabu search, simulated annealing, late-acceptance hill climbing). The greedy heuristic gives an instant baseline, CP-SAT a complete-with-time-budget solver, and Timefold a metaheuristic capable of scaling to large instances. A benchmarking framework with synthetic small, medium, and large datasets (sizes specified in \Cref{sec:evaluation-framework}) ran the same problem instances against all three engines, with engine-version snapshots recorded for reproducibility.
  \item[Why.] Under realistic constraint combinations, the open question is which solver approach performs best on the metrics that operationalise Efficiency in the framework set out in \Cref{sec:evaluation-framework}: scheduled-assignment count, soft-constraint penalty, and runtime within budget. That is a how-not-whether question: independent of whether the visibility gap is real (the interviews surfaced this) or whether HITL is necessary (Admmit fixed this from the project's first week). The multi-engine architecture is the instrument designed to answer that how-not-whether question.
  \item[What happened.] Subprocess plumbing was complex: Timefold runs as a separate Java Virtual Machine (JVM) process, CP-SAT inside a Python process, and the greedy solver inside the TypeScript runtime. Early benchmark results were inconsistent across runs, with the inconsistencies traced to silent solver-library updates between runs rather than to the problem instances themselves.
  \item[Learned.] Reproducibility in a multi-engine architecture must be enforced at the version level: engine-version snapshots recorded per benchmark run removed the silent-update inconsistency. The multi-engine architecture is also a how-not-whether test, not a problem-validation test, and returns at the discussion level under the Efficiency anchor.
  \item[Next.] With the computation infrastructure stable, build the human-in-the-loop interface that would let the coordinator act on the plans the solvers produced.
\end{description}

\subsection{HITL Interface}
\label{sec:iter-hitl}

\begin{description}
  \item[Origin.] Mixed: Admmit-mandate combined with interview-validated direction (the drag-and-drop interaction style was confirmed by coordinator consensus that the system should suggest a plan the coordinator can correct, not replace the coordinator).
  \item[Tried.] A Gantt-style drag-and-drop assignment UI, an explicit inspect / modify / accept / reject control on every algorithm-generated suggestion, and a thin post-hoc constraint check that surfaced basic feasibility violations such as over-booking and missing competence.
  \item[Why.] HITL was Admmit's requirement from project start, and the interviews validated the necessity through a coordinator consensus that automated planning must be correctable rather than authoritative. Without an interactive override path, the coordinator would have authority in principle but no way to exercise it on the timeline they actually work with.
  \item[What happened.] Drag-and-drop worked. State consistency between optimistic UI updates and server mutations occasionally produced stale conflict displays, where the timeline showed a violation the server had already resolved, and required a refresh hook on assignment updates. The thin post-hoc check caught the most common conflicts but was clearly insufficient for the range of edge cases the coordinator would encounter.
  \item[Learned.] The HITL interface needs richer real-time feedback than a thin post-hoc check provides. Coordinator authority over plan content is meaningful only when the system shows what to act on; an empty timeline the coordinator edits blindly delivers Trust/control only on paper.
  \item[Next.] Expand the post-hoc check into a full deviation taxonomy that materialises the constraints the coordinator would otherwise have to remember.
\end{description}

\subsection{Real-time Deviation Detection}
\label{sec:iter-deviations}

\begin{description}
  \item[Origin.] Designer and domain knowledge. The deviation categories extend pain points named in interviews around tacit knowledge and overtime, but the specific category set was not asked about in the interview guide.
  \item[Tried.] A post-generation conflict scanner across multiple deviation categories: overbooking, overtime violations, missing competencies, expired certifications, rest-period violations, vehicle maintenance conflicts, night-work conflicts, and assignments to unavailable employees. Each deviation carries a severity level and a status workflow (open, acknowledged, resolved). Deviations appear inline on the planning timeline and on a dedicated deviations view.
  \item[Why.] A meaningful HITL interface requires the system to show the coordinator what to override toward, not just an empty timeline they edit blindly. The deviation taxonomy is the mechanism through which constraints kept in the coordinator's head become structured conflict data on the timeline.
  \item[What happened.] False positives appeared at edge-case interactions between weekly work schedules and time-off windows: week boundaries and mid-shift absences produced noisy alerts. Rest-period violations were inherently noisy because the regulation has multiple definitions (the forty-eight-hour weekly average, the sixty-hour weekly maximum, the daily thirteen-hour limit), and a deviation flagged under one definition could be compliant under another.
  \item[Learned.] Deviation detection captures tacit knowledge as structured conflict data, but tuning thresholds is itself a coordination problem the system cannot fully resolve. The coordinator decides when a flagged deviation is real, which keeps the override authority where the HITL stance places it.
  \item[Next.] With the HITL interface and deviation feedback in place, expose the user-facing planning-time control through which the coordinator chooses what kind of plan to ask for.
\end{description}

\subsection{User-controlled Plan-time vs Plan-quality Tradeoff}
\label{sec:iter-tradeoff}

\begin{description}
  \item[Origin.] Designer-technical, as a Trust/control refinement that gives the coordinator authority over the planning process itself, not only over plan outcomes.
  \item[Tried.] A coordinator-facing solver-time-budget control. The coordinator selects either \emph{fast / less accurate} (the greedy heuristic with a short time limit) or \emph{slower / more accurate} (the CP-SAT solver with a longer time limit). The time-quality tradeoff is exposed as a deliberate operator choice rather than an internal default.
  \item[Why.] Different planning situations carry different urgency-quality requirements: a sick-leave replanning event under time pressure differs from next week's planning. Coordinator authority over how the plan is generated is a Trust/control dimension distinct from authority over plan content; the override flow handles the latter, the time-budget control adds the former.
  \item[What happened.] The choice itself raised the question of what the user is choosing between. Bare time labels (``ten seconds'' versus ``sixty seconds'') were not legible to a coordinator who does not think in solver-runtime budgets; descriptive labels (``fast suggestion'' versus ``best-effort plan'') were clearer but smuggled in claims about plan quality the system cannot verify per instance. The compromise settled on descriptive labels with a tooltip explaining what the toggle changes.
  \item[Learned.] Exposing a tradeoff in the interface requires the interface itself to explain it, which connects this control to the explanation-as-interface treatment of HITL set out in \Cref{sec:hitl}. The time-budget toggle is a small case of the same principle: the coordinator can act on the tradeoff only when the interface makes the tradeoff legible.
  \item[Next.] Move from operator authority over the planning process to per-company configurability of the planning rules themselves: the soft-constraint weight space.
\end{description}

\subsection{Configurable Soft-constraint Weights}
\label{sec:iter-weights}

\begin{description}
  \item[Origin.] Designer and architectural. The iteration delivers the Adaptability anchor that the project committed to from the start through the Admmit-mandated multi-tenant architecture. The specific weight set was chosen by the designers, not interview-validated as the right set.
  \item[Tried.] A per-company weight configuration UI for soft constraints: workload balance, vehicle preference, transitions between consecutive assignments, priority, and employee preferences. Each weight flows into the CP-SAT objective function so that the same engine produces different plans for different companies without code changes.
  \item[Why.] The interviews confirmed that assignment criteria differ materially across companies, with some weighting workload balance highly and others prioritising vehicle--driver continuity or route familiarity. Adaptability means the same artefact must serve materially different operational rules within a single deployed instance.
  \item[What happened.] User-chosen weight combinations were hard to validate against pathological cases. Weights summing to zero produced unstable rankings, and one weight dominating the others produced plans that satisfied a single criterion at the cost of all others. The user-experience (UX) challenge was explaining what each weight actually means to a coordinator who does not work in optimisation vocabulary.
  \item[Learned.] Configurability for soft weights is a partial realisation of Adaptability. Hard constraints, status workflow definitions, and per-company assignment-rule taxonomies remain hard-coded; broader configurability of those layers is named as a gap in \Cref{sec:limitations} and forwarded to future work in \Cref{sec:future-work}.
  \item[Next.] With eight iterations of artefact in place, the methodological question becomes how to test whether the artefact meets the locked anchors. That question is answered in \Cref{sec:evaluation-framework}.
\end{description}


\section{Evaluation Framework}
\label{sec:evaluation-framework}

The eight iterations in \Cref{sec:iterations} produced an artefact. Testing the claims that artefact embodies needs a framework, set out below. That framework is asymmetric on purpose, following the position carried through \Cref{sec:iter-multiengine}.

RP embodies two claims that the framework does not itself test. The first is that the visibility gap is real, which the seven coordinator interviews surfaced (\Cref{sec:data-collection}). The second is that human-in-the-loop operation is the right operator stance, which Admmit fixed in the project's first week (\Cref{sec:defining-task}). The framework, conversely, tests something RP does not embody as a claim: through the multi-engine architecture and the benchmark protocol described below, it tests how solver approaches compare under realistic constraint combinations, not whether the underlying problem is worth solving. The validate-versus-evaluate distinction that bounds these claims is set out under DSR theory in \Cref{sec:dsr} and is not repeated here; the benchmark is a validation instrument in Wieringa's sense (see \Cref{sec:dsr}), applied at the level of solver behaviour.

This how-not-whether stance carries through to how each anchor is tested, and the coverage is asymmetric. \textbf{Efficiency} is the most directly tested anchor: the multi-engine benchmark is the validation instrument designed for it, comparing scheduled-assignment count, soft-constraint penalty, and runtime-within-budget across the greedy heuristic, OR-Tools CP-SAT, and Timefold metaheuristic on identical problem instances. \textbf{Adaptability} is tested only indirectly: the requirements traceability matrix records that the per-company configurable weight interface (\Cref{sec:iter-weights}) is implemented and verified during development, but no benchmark compares the artefact's behaviour across two genuinely different operational rulebooks. \textbf{Trust/control} is also tested only partially. The inspect / modify / accept / reject flow is implemented, but it has not been user-tested with traffic coordinators, and that gap is forwarded to L8 in \Cref{sec:limitations}.

The benchmark runs against three synthetic datasets whose sizes were chosen to span the fleet ranges named in \Cref{sec:data-collection}. The small dataset (five employees, three vehicles, twenty assignments) is a sanity-scale instance below the smallest interviewed fleet on which every engine should finish in well under a second; the medium dataset (twenty employees, ten vehicles, one hundred assignments) sits at the small-operator end of the interviewed range (around twenty drivers) and exercises the overlap and workload-balance terms without saturating the search; the large dataset (fifty employees, twenty-five vehicles, five hundred assignments) exceeds the largest multi-department fleet observed (around forty-five vehicles) and is a stress-scale instance designed to push CP-SAT toward its sixty-second time budget and probe metaheuristic behaviour. The choice of synthetic over production data follows from two concrete constraints rather than from methodological preference: no transport company has operated the artefact, so production data does not exist for this codebase, and the operational data the seven interviewed companies hold falls under the data-protection bounds the consultation framing in \Cref{sec:data-collection} chose not to cross. The price the framework pays for that choice is that benchmark numbers cannot be projected onto real Norwegian-transport operational data without further validation, a bound named as L5 in \Cref{sec:limitations}.

Alongside the benchmark, the requirements traceability matrix functions as the coverage check rather than as a quality validation. For each functional requirement in the FK-01 to FK-42 range, the matrix records two things: whether it is implemented in the deployed codebase, and whether its behaviour was verified during development. ``Verified during development'' is a narrower claim than passing an automated test suite, and it is the honest one to make: thirty-six of the thirty-seven in-scope functional requirements are implemented and developer-verified, one is partial (FK-30, driving and rest-time status), and the five out-of-scope requirements (FK-38 through FK-42) are documented as future work. The absence of a formal automated test suite is itself a non-functional gap and is named as L9 in \Cref{sec:limitations}; the traceability matrix's role is to make coverage legible, not to substitute for the testing that L9 acknowledges is missing.

Three claims about the artefact are deliberately outside what this framework tests, and each forwards to a specific limitation in \Cref{sec:limitations}. The artefact is not deployed in production, so claims about Efficiency gains under operational time pressure cannot be made from these results: that gap is L6. There is no empirical comparison against Timpex, Opter, or the internal and custom tools the seven interviewed companies use, so claims that RP is better than alternatives in the same market segment are not supported here: that gap is L7. The inspect / modify / accept / reject flow that delivers Trust/control has not been user-tested with traffic coordinators, so claims about its comprehensibility or its behaviour under coordinator workload cannot be made: that gap is L8. The framework that follows in \Cref{sec:validity} addresses the research's epistemic limits rather than the artefact's; this section ends where validation ends, and \Cref{sec:limitations} carries forward what neither addresses.


\section{Validity and Reliability}
\label{sec:validity}

\Cref{sec:evaluation-framework} bounded what the artefact's evaluation can claim; this section bounds what the surrounding research can claim. The framework used is \textcite{malterud2001lancet}'s three standards for qualitative inquiry: relevance (whether the research question is worth answering), validity (whether the study investigates what it claims), and reflexivity (whether the researcher's effect on the process is acknowledged and managed) \parencite[p.~483]{malterud2001lancet}. Each standard binds a different paragraph below.

Relevance and validity bound the interview component first. The seven coordinators were a purposive sample matched to the research question (coordinator experience of the visibility gap, not its prevalence in Norwegian transport), and \textcite[p.~485]{malterud2001lancet} treats purposive sampling as the appropriate strategy for qualitative inquiry. Validity within that sample rests on triangulation across the seven calls: the visibility gap and the operator-vs-owner asymmetry surfaced across companies of different size and system maturity rather than from any single coordinator. The findings transfer in Malterud's bounded sense of applicability within a specified setting rather than to the population at large \parencite[p.~486]{malterud2001lancet}, and the bounds are named as L1 (sample size, brief duration) and L2 (self-selection from Admmit's customer roster) in \Cref{sec:limitations}.

Validity for the system component is narrower. Two instruments carry it, both set out in \Cref{sec:evaluation-framework}: the requirements traceability matrix that records implementation and developer-verified behaviour for each functional requirement, and the multi-engine benchmark that compares the three engines on identical synthetic instances. Together they validate the artefact in Wieringa's sense (predicting how it would behave once deployed) rather than evaluating it in deployment (see \Cref{sec:dsr}). The gaps that bound this validation (no production deployment, no user testing of the override flow, no automated test suite) are forwarded as L6, L8, and L9 in \Cref{sec:limitations}.

Reflexivity addresses the researcher configuration directly. The research team and the development team are the same two people, and one author has a prior affiliation with Admmit that predates the project. \textcite[p.~484]{malterud2001lancet} treats preconceptions as separate from bias unless the researcher fails to disclose them; declaring the dev-team-as-research-team configuration and the Admmit affiliation keeps reflexivity intact, and failing to declare them would be what breaks it. Both the prior expectation that a visibility gap might exist and the prior commitment to HITL are named in \Cref{sec:defining-task,sec:dsr-dsrm} for the same reason. The researcher--developer overlap and the author affiliation with Admmit are forwarded together as L3 in \Cref{sec:limitations}.
