\chapter{Conclusion}
\label{chap:conclusion}


\section{Summary}
\label{sec:summary}

Norwegian transport companies operate without systematic visibility into resource utilization. Across the seven interviewed companies, overtime is handled reactively, idle time between assignments is invisible in current tools, and load balancing across drivers depends on coordinator memory. The demand to address that gap is articulated by owners and by Admmit, not by the coordinators who must run any algorithm-assisted system day to day, and the same configuration mirrors Bainbridge's \emph{Ironies of Automation}.

This thesis used Design Science Research as its paradigm, applying the Peffers DSRM activities through eight named iterations. The empirical foundation was built from semi-structured interviews with seven traffic coordinators and transport managers across Norway, analysed through Braun and Clarke's six-phase thematic analysis. The artefact, RP, is a Next.js + tRPC + PostgreSQL platform with a multi-engine solver layer (greedy heuristic, OR-Tools CP-SAT, Timefold metaheuristic) and a drag-and-drop human-in-the-loop timeline. The evaluation framework benchmarks the three engines against three synthetic datasets and complements the benchmark with a requirements traceability matrix.

The interviews surfaced the resource-utilization visibility gap as the primary finding and the operator-vs-owner asymmetry as the configuration that makes the gap a finding worth surfacing. Forty-two functional requirements were prioritised under MoSCoW, of which thirty-seven are in scope and thirty-six are implemented and developer-verified. The fit/gap analysis produced five fit items where Timpex, Opter, and the internal or manual tools the seven companies use already meet a need, and fourteen gap items where they do not. The multi-engine benchmark documents a seventy-two per cent canonical-penalty reduction in the Timefold engine after the constraint-provider rework on 2026-04-23. Twelve hierarchical limitations (L1 through L12) bound what the thesis can claim, the most consequential being the absence of production deployment (L6) and of user testing with coordinators (L8). The thesis's central qualification is that RP supports but does not replace coordinator judgement, and that its argument applies under the stakeholder asymmetry that runs across the seven interviewed companies, not in low-variation fleets where utilisation patterns are uniform across drivers.


\section{Answers to Research Questions}
\label{sec:answers}

\begin{quote}
How can an algorithm-assisted resource planning platform improve resource utilization in Norwegian transport companies (reducing overtime, idle time between assignments, and uneven driver load) while remaining accountable to the traffic coordinator who operates it?
\end{quote}

An algorithm-assisted resource planning platform can improve resource utilization across the three named dimensions when it makes utilisation visible before any algorithm reorders it (\textbf{Efficiency}), keeps the coordinator's authority over every algorithm-generated assignment through inspect, modify, accept, or reject (\textbf{Trust/control}), and configures its soft-constraint weights per company so that materially different operational rulebooks are served without code changes (\textbf{Adaptability}). RP demonstrates this configuration through a planning timeline that exposes utilisation patterns that current systems leave invisible, a four-layer HITL interface (deviation viewer, score panel, time/quality control, drag-and-drop override) that holds the coordinator's authority structurally rather than as a usability feature, and per-company configurable soft-constraint weights that flow through a shared CP-SAT objective function. The evidence base is bounded by the absence of production deployment (L6), the absence of user testing with coordinators (L8), and the use of synthetic rather than production benchmark data (L5); the artefact is validated in Wieringa's sense (predicting how it would behave once deployed), not evaluated in deployment.

\begin{quote}
\textbf{SQ1} — How is resource-utilization visibility distributed across operator and owner stakeholders in current Norwegian transport-company planning practice?
\end{quote}

Resource-utilization visibility is distributed asymmetrically. Across the seven interviewed Norwegian transport companies, owners and Admmit articulate utilisation visibility and the demand for an algorithm-assisted plan; coordinators articulate speed, override authority, and a system that does not get in their way. The asymmetry is not a peripheral observation: it is the configuration that makes the visibility gap a finding worth surfacing under \textbf{Efficiency} (§5.1.1), and it is the configuration Bainbridge's \emph{Ironies of Automation} anticipates. The answer is bounded by L1 (sample size of seven), L2 (Admmit-customer recruitment), L3 (author affiliation with Admmit), and L4 (interview-guide coverage gap), so the thesis claims the asymmetry as the consistent configuration in the consultation pool rather than as a population-level statistic for Norwegian transport.

\begin{quote}
\textbf{SQ2} — How can an algorithm-assisted planning system be designed so that automated optimisation improves resource utilization while remaining operable and overridable by the traffic coordinator?
\end{quote}

An algorithm-assisted planning system can stay operable and overridable by the coordinator when override is structural rather than optional, when explanation is part of the interface rather than added afterwards, when trust calibration is a design target rather than an accident of deployment, and when the system stops at Parasuraman level 5 (the coordinator must approve, not merely fail to object) rather than at level 6. RP delivers these conditions through the planning timeline (the suggested plan is shown in the same Gantt view used for manual work), the deviation viewer (each conflict surfaces type, affected assignment, and affected resource), the post-solve summary panel (solution score and counts of hard and soft violations), the time/quality control (fast suggestion versus best-effort plan with explanatory tooltip), and drag-and-drop override that saves with a warning rather than blocks (§5.1.2, \textbf{Trust/control}). The answer is bounded by L8 (no user testing with coordinators) and L10 (HITL as Admmit mandate, not validated level), so the design argument is design plus theory rather than design plus theory plus operational evidence.

\begin{quote}
\textbf{SQ3} — To what extent and under what limitations does Ressursplanlegger demonstrate measurable improvement in resource utilization?
\end{quote}

RP demonstrates measurable improvement at the level of solver behaviour on synthetic instances, not at the level of operational utilisation in deployed Norwegian transport. The multi-engine benchmark in §4.5 documents a seventy-two per cent reduction in canonical penalty for the Timefold engine after the 2026-04-23 constraint-provider rework, OR-Tools CP-SAT as the strongest individual engine on small and medium instances, and the operational scaling boundary at the five-hundred-assignment large instance under a sixty-second time budget. The requirements traceability matrix records thirty-six implemented-and-developer-verified requirements out of thirty-seven in scope. Improvement at the level of overtime, idle time, and load balance in real fleets is not demonstrated by these results and cannot be without the production pilot named in §6.3 (\textbf{Adaptability}, with cross-anchor binding to \textbf{Efficiency} and \textbf{Trust/control}). The answer is bounded by L5 (synthetic benchmarks rather than production data), L6 (no real-world deployment), L7 (no empirical comparison against existing TMS), L9 (algorithm evaluation against own benchmarks only), L11 (single domain), and L12 (boundary cases described, not quantified).

The contribution of the thesis sits at three levels. The artefact (RP) is a deployable algorithm-assisted planning platform that fills the planning gap existing Transport Management Systems leave open. The empirical foundation is a structured account of the visibility gap and the operator-vs-owner asymmetry in seven Norwegian transport companies, together with a forty-two-requirement register and a fit/gap analysis grounded in those interviews. The design knowledge is the multi-engine architecture's how-not-whether validation framing, the four-layer HITL interface treatment of Trust/control, and the configurable soft-constraint weight mechanism for Adaptability inside a multi-tenant deployment.


\section{Future Work}
\label{sec:future-work}

Future work is grounded in specific named limitations from §5.4. Generic items ("further research could investigate...") are not on this list.

A driver-facing page within the web application is the highest-priority extension. The interviews surfaced it as a recurring request across companies, and its absence interacts with L8: a coordinator-driven flow that has not been user-tested with coordinators is doubly bounded when the driver-facing endpoint is also missing. Implementing the driver page and conducting coordinator user testing on the override flow at the same time would close part of L8 directly.

Billing and invoicing integration was the most frequently named missing feature across the seven interviews. It addresses the adoption barrier identified in §5.2: a transport company that adopts RP keeps its existing TMS for invoicing, and an integration that lets the two systems exchange data through a documented interface would lower the adoption threshold meaningfully. The integration sits in the order-management category that Timpex and Opter already cover, so the integration project is bridging rather than replacing.

A production pilot with one of the seven interviewed companies is the single most valuable methodological move available to a follow-on project. It addresses L5 (synthetic benchmarks rather than production data) and L6 (no real-world deployment) at the same time, lets the artefact's Efficiency claims be tested against operational utilization data rather than against synthetic instances, and would generate the empirical comparison that L7 currently bounds.

Real-time replanning for sick-leave events is a feature several interviewees flagged. Norlog Tana described sick-leave as the dominant operational problem of the day, and Ottem reported a structurally similar problem from order cancellations. A re-solve flow that takes a partial plan plus a disruption (a driver going sick, an order being cancelled) and returns an updated plan, with the coordinator's previous decisions preserved where possible, would address part of L9 (algorithm evaluation against own benchmarks only) by introducing a real-event-driven evaluation regime.

Algorithm scaling improvements address the operational scaling boundary documented in §4.5 and forwarded as part of L9. OR-Tools CP-SAT does not complete the five-hundred-assignment large instance within sixty seconds; longer time budgets, decomposition strategies, or a warm-start from the greedy heuristic could shift the boundary. The Timefold lex-tier behaviour on the large instance (better on legal, business-critical, and preference tiers but losing on hard) is itself a design question for future work: tuning the constraint provider to reduce the hard-tier delta without sacrificing the medium and soft tiers would let the portfolio engine select Timefold on instances where lex currently selects greedy.

Broader Adaptability addresses the partial realisation named in §3.5.8 and L4. Hard-constraint configurability, status-workflow definitions per company, and per-company assignment-rule taxonomies are all hard-coded in the current artefact. A subsequent iteration that opens these layers to per-tenant configuration, with an interview-guide pass that probes the rule differences L4 names, would deliver a fuller realisation of Adaptability.

Cross-domain replication addresses L11. The visibility gap, the operator-vs-owner asymmetry, the multi-engine architecture, and the configurability mechanism are all anchored in Norwegian transport in this thesis. Whether they transfer to nurse rostering, airline crew scheduling, or other Nordic transport markets is a question for future replication. The structural argument (multi-resource assignment plus tacit knowledge plus stakeholder asymmetry) is general enough to suggest transferability, but transferability is not entailed by the structural argument and would have to be tested.

The closing claim of this thesis is not about RP alone. Algorithm-assisted planning under stakeholder asymmetry in Norwegian transport is feasible, and the conditions under which it is feasible are structural rather than incidental: the operator must keep authority over every algorithm-generated assignment, the system must explain itself in the operator's vocabulary rather than in solver internals, and the configuration that distinguishes one company from another must live as configuration rather than as code. RP demonstrates these conditions inside a single bachelor-thesis artefact. Whether the same conditions hold in adjacent domains, and whether they hold under operational time pressure with real coordinators on real days, is the empirical work this thesis hands to the next one.
