\chapter{Introduction}
\label{chap:introduction}

% See context/outline.md Ch 1 for paragraph-level plan.

\section{Background and Motivation}
\label{sec:background}

Norwegian transport companies move large volumes of goods every day. In 2025, these companies carried about 252 million tonnes on behalf of their customers, almost all of it within Norway \parencite{ssb2026godstransport}. Behind every load, in every transport company, a person called the traffic coordinator decides who drives what, in which vehicle, and when. That decision is repeated dozens of times a day, and how well it lands shows up in three concrete places: how much overtime drivers run, how long they wait between assignments, and how evenly the work is shared across the team. To understand how this decision is made in current practice, the project drew on interviews with traffic coordinators in Norwegian transport companies and on an ongoing dialogue with Admmit, the Norwegian software firm that offered this thesis as a bachelor task. Across those conversations, the same pattern surfaced. The information needed to track overtime, idle time, and load balance is in the company's systems, but it is spread across an order tool, a timesheet, the coordinator's spreadsheets, and the coordinator's memory, and it is hard to pull together into a single working overview. The coordinator can answer any specific question about a specific driver but cannot easily step back and see how the week as a whole is shaping up for the team. The thesis takes this as its starting point: the data needed to plan well already sits in the company's systems, but in its current form it is hard for any one person to read at a glance.

The demand for a system that helps with this planning has more than one origin. Admmit, the Norwegian software firm that offered this thesis as a bachelor task, has worked with transport-company owners for years and reports a recurring request from them: an automated planning tool that the coordinator can supervise, on the grounds that better-balanced plans cut overtime and idle hours and raise the company's earnings. Admmit translated that owner-side demand into two requirements for any such system. The first is a Human-in-the-Loop (HITL) operating model, so the coordinator can inspect, modify, accept, or reject every algorithm-generated assignment before it takes effect. The second is a multi-tenant architecture, in which a single deployed system serves several customer companies at once with their data and rules kept separate. The interviews with traffic coordinators were carried out alongside the dialogue with Admmit and put those two requirements in front of the people who would actually run the system. Coordinators confirmed that the requirements made sense and added wishes of their own, most often a single-view overview of utilisation across drivers and vehicles, not only for the day ahead but for the week and the longer rotations that follow. Several coordinators said directly that they would welcome an algorithm proposing the day's assignments. Others were more cautious, pointing to the unpredictability of transport work as a reason to keep human judgement at the centre of every decision. No coordinator opposed an algorithm proposing assignments; what each insisted on was the right to override. The split between the firm and the owners articulating the demand for automation and the coordinators who must operate the result day to day is the structural pattern \textcite{bainbridge1983ironies} describes in \emph{Ironies of Automation}, and it runs across the empirical foundation of this thesis.

Current planning practice in Norwegian transport companies is held in place by two things: the knowledge that sits in the coordinator's head, and the software that surrounds it. Coordinators carry much of what is needed to plan well in their heads, who holds the right licence for which load, who knows the difficult routes, who is close to the overtime ceiling, which vehicle suits which job. This kind of knowledge, learned on the job and rarely written down, is sometimes called tacit knowledge, and it works only as long as the same coordinator stays in the role. The software side does not pick up the slack. The two commercial Norwegian Transport Management Systems (TMS) named by interviewees, Timpex and Opter, handle order management and invoicing well, but neither generates assignment plans on its own, and Timpex is described as slow under heavy use. The other interviewed companies plan from internal tools, spreadsheets, memory, or the phone. None of the systems in current use takes on the assignment step itself: deciding which driver and vehicle perform which order.

Ressursplanlegger (hereafter RP) is the artefact this thesis builds and evaluates. RP is an algorithm-assisted planning platform: an algorithm proposes a plan for a given day, week, or longer horizon, and a drag-and-drop timeline lets the coordinator inspect, modify, accept, or reject every algorithm-generated assignment before it takes effect. Three different optimisation engines sit behind the proposal: a fast greedy heuristic, Google's OR-Tools CP-SAT, and the Timefold metaheuristic. The same artefact can be tuned per customer company through configurable preference weights, without code changes. RP is not a Transport Management System in the sense Timpex and Opter are: it does not handle order management or invoicing. It addresses the planning step those systems leave open.


\section{Anchor Concepts}
\label{sec:anchors}

In the dialogue with Admmit and in the interviews with traffic coordinators, three things stood out as what the system had to deliver. They show up again and again in the rest of the thesis. Every feature, every finding, and every limitation in the chapters that follow comes back to one of them: Efficiency, Trust/control, and Adaptability.

\subsection{Efficiency}
\label{sec:anchor-efficiency}

The system has to help the company make better use of its drivers and vehicles. Three things should improve: drivers should run less overtime, they should spend less time waiting between assignments, and the work should be shared more evenly across the team. Before the system can improve any of these, it first has to make them easy to see for the coordinator and for the owner. The data is already in the company, but no one can read it at a glance today.

\subsection{Trust/control}
\label{sec:anchor-trust-control}

The coordinator has to stay in charge of every assignment the algorithm proposes. For each one, the coordinator should be able to look at it, change it, accept it, or reject it before it takes effect. Trust in the algorithm grows over time, but only if the coordinator can see what it does and change it when needed. A system that made the plans on its own and left the coordinator with no way to step in would not work in the kind of company this thesis is built for.

\subsection{Adaptability}
\label{sec:anchor-adaptability}

The same system has to work for very different companies without being rebuilt each time. Transport companies vary a lot: a small operator with around ten drivers does not run the same way as a firm with forty-five vehicles split across three departments. The system has to fit both, with the rules and priorities adjusted per company in the settings rather than in the code itself.


\section{Research Questions}
\label{sec:research-questions}

\begin{quote}
How can an algorithm-assisted resource planning platform improve resource utilization in Norwegian transport companies (reducing overtime, idle time between assignments, and uneven driver load) while remaining accountable to the traffic coordinator who operates it?
\end{quote}

The main question ties the two things the system has to deliver: a real improvement in how the company uses its drivers and vehicles, and a way of working that keeps the coordinator in charge of the result. A system that improved the numbers but took the decision away from the coordinator would not answer the question, and a system that left the coordinator fully in charge but did not improve anything would not answer it either.

Three sub-questions break the main question into smaller pieces that can be answered later in the thesis.

\begin{quote}
\textbf{SQ1} — How is resource-utilization visibility distributed across operator and owner stakeholders in current Norwegian transport-company planning practice?
\end{quote}

SQ1 is about the situation as it stands today. It connects to \textbf{Efficiency} and is answered in §5.1.1, drawing on the interviews and the gap they show between what owners and coordinators can see in current planning.

\begin{quote}
\textbf{SQ2} — How can an algorithm-assisted planning system be designed so that automated optimisation improves resource utilization while remaining operable and overridable by the traffic coordinator?
\end{quote}

SQ2 is about how the system should be designed. It connects to \textbf{Trust/control} and is answered in §5.1.2 by applying human-in-the-loop theory to the parts of the artefact that handle overrides, conflicts, scoring, and the trade-off between speed and plan quality.

\begin{quote}
\textbf{SQ3} — To what extent and under what limitations does Ressursplanlegger demonstrate measurable improvement in resource utilization?
\end{quote}

SQ3 is about how well the system actually works. It connects mainly to \textbf{Adaptability}, with links to the other two anchors as well, and is answered in §5.1.3 and §5.4. The answer draws on the benchmark of the three solver engines, the check of how each requirement was met, and the twelve limitations that mark out what the thesis can and cannot claim.


\section{Scope and Delimitations}
\label{sec:scope}

The system the thesis builds covers the planning step itself: turning a list of orders for a given period into a concrete plan that says who drives what, in which vehicle, and when. The length of that period is set by the coordinator in the dashboard and ranges from a single day to a week or longer, since the companies the system has to fit do not all plan on the same horizon. Around that core, the system holds the information the planning step needs: who the drivers are and what they are qualified to do, what vehicles the company has and what each one can carry, when people are working and when they are not. Three different solver engines run over that information to produce a proposed plan, the coordinator can weight priorities differently per company, and conflicts are flagged when something does not fit. A dashboard summarises how the team is doing on the three anchors. Each of these parts is in scope because it serves at least one anchor: making utilisation visible and improving it under Efficiency, supporting override and explanation under Trust/control, and supporting per-company tuning under Adaptability.

Several things the system deliberately does not do are worth naming, because they came up in the conversations and could otherwise be assumed. The system is not a driver-facing app and does not push notifications to drivers: it is a planning tool, not a dispatch tool. It does not handle billing or invoicing, which is what Timpex and Opter already cover, and bridging the two is a project of its own. It covers Norwegian transport only, since the regulatory frame is national. It does not track vehicles in real time on the road; it plans the day, it does not monitor the day's execution. It does not try to plan the order in which a single driver should visit a list of stops, sometimes called the Vehicle Routing Problem, since the question it answers is who does what, not in which order. And it does not let the algorithm run without a human in the loop; that is a deliberate Admmit requirement, not a setting that can be turned off.
