\chapter{Findings}
\label{chap:findings}

This chapter reports what the seven interviews surfaced and what was built, without interpretation. \Cref{sec:interview-findings} reports the themes the thematic analysis named. \Cref{sec:requirements} lists the functional and non-functional requirements derived from those themes. \Cref{sec:fitgap} compares what existing systems cover with what is missing. \Cref{sec:system-description} describes Ressursplanlegger as it actually is: architecture, features, the user interface, and the technology stack. \Cref{sec:algorithm} reports the optimisation algorithm and the benchmark results that test it. \Cref{sec:dsr-artifacts} maps each project artefact to its DSR category. \Cref{sec:process-documentation} documents the project process. Interpretation, anchor binding, and limitation analysis belong to Chapter 5.


\section{Interview Findings}
\label{sec:interview-findings}

The seven interviewed companies plan their day's transport assignments manually. Bergen Bulk Transport plans by phone and from memory, Harlem Solutions plans inside Opter, three Norlog branches plan inside Timpex, Ottem plans inside an internal order system across three departments, and Nordic Crane plans inside its own system across crane and transport divisions. None of these systems generates assignment plans automatically. In every interviewed company, a traffic coordinator reads the day's orders, looks at who is available, and writes the assignments down by hand. The scale runs from a small operator with eight to twenty drivers (Bergen Bulk Transport) up to a multi-department fleet of around forty-five vehicles across three departments (Ottem). Despite this range, the manual matching step is the same.

The primary surprising theme is what the analysis calls the \textbf{resource-utilization visibility gap}. Across the seven calls, coordinators consistently reported that they handle overtime reactively rather than systematically, that idle time between assignments is invisible in their current tools, and that load balancing across drivers depends on memory rather than on any structured overview. Coordinators do not, on their own, articulate utilization optimisation as something they need. The articulation comes from Admmit and from the business side. This is the second surprising theme: an \textbf{operator-vs-owner asymmetry} that runs across the seven interviews. Owners and Admmit ask for utilisation visibility and for an algorithm-assisted plan. Coordinators ask for speed, override authority, and a system that does not get in their way. Both themes are central to the discussion in §5.1.1.

Pain points named in the interviews, ranked by frequency across the seven calls, are slow legacy software, tacit-knowledge dependency, and lack of capacity overview. Slow legacy software is named most often. Norlog Mo i Rana describes Timpex as "ekstremt tregt" with a "gammelt grensesnitt" and says speed is "100 prosent" the most important property to improve. Norlog Lakselv confirms that Timpex performance falls dramatically when more users log in concurrently. Tacit-knowledge dependency is the second most frequent theme: every interviewed coordinator holds operational knowledge that is not formally captured by their current system, including who has the right licence and competence, who is experienced on which routes, who is approaching overtime limits, and which vehicle suits which job. Lack of capacity overview comes third: no interviewed company has a single view of driver availability, vehicle status, and assignment load.

Sick-leave handling varies materially across the interviewed companies. Some coordinators describe it as trivial: Norlog Lakselv handles a sick-day with "two or three keystrokes", and Nordic Crane reports completing a replanning event in about twenty-seven seconds. Others describe sick-leave as the dominant operational problem of the day. Norlog Tana states it explicitly: "jeg sitter jo egentlig med den problemløsningen hver eneste dag" (I sit with that problem-solving every single day). The difference depends on how many drivers are available, how flexible the routes are, and whether the routes require special competence such as scanning at Norlog Tana. Ottem reports a structurally similar problem from the order side rather than the driver side: orders are cancelled unpredictably and require similar reshuffling.

Attitudes toward automation are split rather than aligned. Norlog Lakselv, Norlog Mo i Rana, and Norlog Tana are positive toward algorithm-assisted planning. Harlem Solutions, Ottem, and Bergen Bulk Transport are sceptical, primarily because transport is unpredictable and they want to keep human judgement in the loop. Nordic Crane is partially positive and asks specifically for automatic distribution eighteen to twenty-four hours before an assignment runs, rather than full real-time automation. Even the sceptical interviewees accept the idea of an algorithmic suggestion. Ottem puts the consensus most clearly: "det må være noen som sitter og tenker bak meg da" (there must be someone sitting and thinking behind me). The point of agreement across all seven calls is that the system should propose a plan the coordinator can correct, not replace the coordinator.

When asked what they consider when matching a driver to an assignment, the interviewed coordinators name the same set of criteria, in the following frequency-ranked order: availability and position (who is free and nearest), experience (especially on demanding routes), driving and rest-time status (legally required), competence and licence (matching the certificate to the assignment), route familiarity (especially in northern Norway), equipment match (the right vehicle for the job), and overtime status (taken into account particularly toward annual limits). The list is consistent across companies of different size and current system, but the relative weight each company places on each criterion differs. Some companies prioritise workload balance, others prioritise vehicle--driver continuity or route familiarity. The variation across companies motivates configurable soft-constraint weights in the artefact.


\section{Requirements}
\label{sec:requirements}

The themes named in \Cref{sec:interview-findings} were translated into forty-two functional requirements (FK-01 through FK-42) and sixteen non-functional requirements (IFK-01 through IFK-16), prioritised under MoSCoW. The visibility-gap theme produced the timeline view (FK-01) and the conflict-detection requirements (FK-24 through FK-30) at Must. The assignment-criteria theme produced the hard-constraint requirements (FK-19) at Must and the soft-constraint requirements (FK-20 through FK-23) at Should. The automation-attitudes theme produced the override (FK-04) and weight-configuration (FK-21) requirements at Must and Should respectively. Recurring assignments (FK-05) came from Norlog Lakselv and Norlog Mo i Rana at Should. Multi-tenancy with role-based invitations (FK-31 through FK-34) came from Admmit's mandate at Must. Five requirements (FK-38 through FK-42) are in the register as Won't for this version: a driver-facing page, billing integration, push notifications to drivers, an admin panel, and GPS tracking. They are documented as future work, named in interviews but out of the thesis scope.

The non-functional requirements set targets the artefact must meet rather than features it must implement. Algorithm response is bounded at under three seconds for the greedy solver on up to one hundred assignments (IFK-01) and at under sixty seconds for OR-Tools CP-SAT on up to five hundred assignments under a configurable time limit (IFK-02). API response is bounded at under one second per request under normal load (IFK-03). Initial page load on a standard broadband connection is bounded at under three seconds (IFK-04). Five to fifty concurrent coordinators per company are supported without degradation under serverless auto-scaling (IFK-05). Time to learn the core planning workflow is bounded at under thirty minutes for a coordinator familiar with transport planning (IFK-06). Service availability is bounded at ninety-nine per cent or higher (IFK-07). Authentication is session-based via Better Auth, sessions stored in PostgreSQL with expiry (IFK-08). Authorisation scopes all data access to the authenticated user's \texttt{companyId} (IFK-09). All API inputs are validated with Zod schemas (IFK-10). SQL injection is prevented by Prisma ORM's parameterised queries throughout (IFK-11). The full register, with sources mapped to interviews and implementation status verified against the codebase, is held in the project's working notes alongside the requirements traceability matrix.


\section{Fit/Gap Analysis}
\label{sec:fitgap}

A side-by-side comparison of what existing systems cover and what they do not produced five fit items and fourteen gap items. The fit items are order and assignment registration, invoicing and billing, driver notification, basic driver and vehicle data storage, and assignment status tracking for invoicing purposes. Both Timpex and Opter cover order registration and invoicing fully; both have driver-facing notification apps (Timpex Confirm and Opter's order app); both store basic driver and vehicle data; both track assignment status for invoicing. Custom systems at Ottem and Nordic Crane cover similar ground, tailored to each company. Manual planning at Bergen Bulk Transport covers nothing systematically: orders live on phones and in coordinator memory.

The fourteen gap items concentrate on the planning step itself. None of the existing systems provides a unified daily planning view across assignments, drivers, and vehicles. None generates an assignment plan automatically. None checks for scheduling conflicts. None provides a driver-availability overview that reflects work schedules and time-off in one place. Competency and certification tracking with expiry dates is not provided by Timpex or Opter, even though the interviewed companies need it. A capacity overview that answers the question "is the day covered?" is absent across all existing systems. Manual override of an algorithmic suggestion is not applicable to systems that produce no suggestions, but the requirement was stated clearly across the interviews. Recurring assignments are unsupported as first-class entities; coordinators copy and re-enter recurring jobs by hand. Sick-leave handling does not link absence to assignment coverage in any existing system. Driving and rest-time monitoring is not integrated with planning. Vehicle maintenance tracking, where it exists at all, lives outside the TMS. Utilisation metrics for drivers and vehicles are not provided. Multi-tenant data isolation across companies or departments, mentioned by Ottem, is not standard in single-company TMS installations. System performance under concurrent users is the fourteenth gap, named by every Timpex user as a major pain point. The gaps cluster: order management and billing are well-covered, the assignment decision and what depends on it are not.

Vendor-specific factual context fills out the picture. Timpex is described as extremely slow under concurrent use and is used for order management, invoicing, and driver notification across the three Norlog branches. Opter is used by Harlem Solutions for order management and invoicing, with a driver-facing order app; it is described primarily as a billing tool with no planning support. Custom systems at Ottem and Nordic Crane are tailored to each company's operations and do well at what they were built for, but offer no standardisation, no optimisation, and are difficult to maintain. Bergen Bulk Transport plans by phone and from memory, with no software in the planning loop. The core gap, in one sentence, is that all existing systems address what happens after the assignment is made (order registration, invoicing, delivery tracking) and none addresses the assignment decision itself: who drives what, with which vehicle, and why. RP occupies that gap. It is a planning and decision-support tool, not an order-management or invoicing tool, and would deploy alongside Timpex or Opter rather than replace them.


\section{System Description}
\label{sec:system-description}

RP is a full-stack monorepo built on Next.js 15 with the App Router, with React 19 client components and React Server Components, TypeScript end-to-end, and tRPC 11 as the typed remote-procedure-call layer between the browser and the Next.js server. There is no separate backend service for standard operations: the Next.js API routes serve as both the HTTP server and the business-logic layer. The browser communicates with the server via tRPC over HTTPS, with TanStack Query handling client-side server-state caching. The server uses Better Auth for session management (sessions persisted to PostgreSQL with expiry) and Prisma 7 as the ORM against PostgreSQL. Three optimisation engines run as subprocesses invoked by the Node.js server: a greedy solver (Python, custom code with no dependencies), an OR-Tools CP-SAT solver (Python, Google's open-source constraint-programming engine), and Timefold (Java, an open-source metaheuristic constraint-solver platform). Each engine receives a problem description on standard input and returns a solution on standard output. Multi-tenant data isolation is enforced at two layers: every entity in the database carries a \texttt{companyId} foreign key, and every tRPC procedure resolves the authenticated user's company from the session before querying.

The deployed feature inventory matches the in-scope list. The planning timeline at \texttt{/ressursplanlegger} shows all assignments, drivers, and vehicles for a selected day in a Gantt-style view, with vertical stacking of overlapping assignments and resource-utilisation indicators rendered inline. The coordinator can manually assign a driver and a vehicle to an assignment, override any algorithm-generated assignment by drag-and-drop, set priority levels, and approve the daily plan. Recurring assignments (daily, weekly, biweekly, monthly) are first-class entities. The assignment status workflow runs from planned through approved, in-progress, completed, and cancelled. Driver and vehicle management cover competencies and certifications with expiry tracking, weekly work schedules, time-off, vehicle types and capacity, vehicle maintenance dates, status flags, and trailers. Multi-tenancy is delivered through company-scoped workspaces with email-and-password authentication via Better Auth, role-based invitations (owner, admin, member), and session-based access control. The dashboard at \texttt{/oversikt} aggregates utilisation metrics and key operational indicators. The deviation viewer at \texttt{/avvik} lists each active conflict by type, affected assignment, and affected resource.

The user interface is built around the planning timeline and the coordinator's three repeated actions: read the day, decide what to change, and approve. The coordinator opens \texttt{/ressursplanlegger} to see the day's plan, either as the algorithm proposed it or as it stood after the previous edit. Drag-and-drop on the timeline reassigns an assignment to a different employee row; conflicting changes save with a warning rather than a block. Modal dialogs handle creating, editing, and deleting assignments, employees, and vehicles. The solver is invoked from a button on the planning timeline; before running, the coordinator chooses between fast / less accurate and slower / more accurate, which selects the greedy heuristic with a short time limit or OR-Tools CP-SAT with a longer time limit. After a solver run, a summary panel reports the solution score and counts of hard and soft violations. The deviation viewer at \texttt{/avvik} provides a dedicated screen for working through conflicts. The dashboard at \texttt{/oversikt}, the all-assignments list at \texttt{/alle-oppdrag}, and the employee and vehicle management screens at \texttt{/ansatte} and \texttt{/kjoretoy} support the coordinator's adjacent tasks. User and company settings live at \texttt{/innstillinger}.

The technology stack is deliberately conservative on the application side and pluggable on the algorithm side. Next.js 15 with App Router was chosen because it provides server-side rendering, file-based routing, and an integrated API layer in one framework, removing the need for a separate backend service. tRPC 11 was chosen because it makes every API call type-checked at compile time across client and server, which removes a class of bugs that would otherwise need run-time tests. Prisma 7 was chosen because it generates typed database queries from the schema, prevents SQL injection through parameterised queries by construction, and supports multi-tenant scoping cleanly through query filters. PostgreSQL was chosen as the database because it handles the relational data model well and is supported as a managed service by the chosen hosting platform. Better Auth was chosen for authentication because it integrates with Next.js Server Components and supports session-based access control with token-based company invitations. Tailwind CSS v4 with shadcn/ui and Radix UI was chosen for styling and components because the combination supports rapid UI work without giving up control over visual detail. The three optimisation engines were chosen because each occupies a different point on the speed-quality tradeoff, and a pluggable subprocess architecture allows the solver written in the language best suited to it to run independently of the application runtime: Python for the greedy baseline and for OR-Tools, Java for Timefold.


\section{Optimisation Algorithm}
\label{sec:algorithm}

The multi-engine architecture tests how solver approaches compare under realistic constraint combinations. It does not test whether the visibility gap is real (the interviews surfaced this) or whether human-in-the-loop operation is necessary (Admmit fixed this from project start). The benchmark is therefore a methodologically independent test of how, not of whether: it answers which solver approach performs best on the metrics that operationalise \textbf{Efficiency} on identical synthetic problem instances. The framing repeats §3.5.4, §3.6, and reappears in §5.1.1.

The problem the algorithm solves is an assignment problem with two paired resources per task. The input is a set of assignments for one operational day, each with a fixed time window, a vehicle-type requirement, an optional competency requirement, and a priority level; a set of employees with competencies, certifications with expiry dates, weekly work schedules, time-off windows, and an annual hour budget; and a set of vehicles with type, capacity, and current status. The output is a set of (assignment, employee, vehicle) triples, with explicit handling for assignments that cannot be feasibly placed (returned as unassigned with a structured reason). Decision variables are binary: does employee $e$ and vehicle $v$ jointly cover assignment $a$? The algorithm chooses an assignment of employees and vehicles to the day's transport tasks that respects the hard constraints and minimises a weighted soft-constraint penalty.

Three solver approaches are integrated in a pluggable solver registry. The greedy heuristic walks the priority-sorted assignment list once and gives each assignment the first feasible (employee, vehicle) combination found. It is fast, milliseconds per instance, and produces a feasible plan, but it commits early to locally good choices that can block later high-priority assignments. The OR-Tools CP-SAT solver formalises the problem as constraint-programming variables and constraints, searches the feasible space using propagation and backtracking, and returns near-optimal plans within a configurable wall-clock budget (sixty seconds in the benchmark configuration). It dominates on small and medium instances but cannot complete the large instance within sixty seconds. Timefold is a metaheuristic engine running tabu search, simulated annealing, and late-acceptance hill climbing under Timefold's \texttt{HardMediumSoftScore} model, with per-slot \texttt{feasibleEmployees} and \texttt{feasibleVehicles} value-range providers, soft weights wired through a \texttt{SoftWeights} problem fact, sum-of-squares workload balance, and dedicated unassigned-slot constraints in the medium tier so that coverage is rewarded separately from feasibility. Each engine reads the same shared constraint configuration, so comparisons measure search behaviour, not constraint definitions. A portfolio engine ("auto") runs all three in parallel and selects the lexicographically best candidate across the four scoring tiers (hard, legal, business-critical, preference), re-scoring every child candidate with the canonical scorer before lex comparison.

Six hard constraints must be satisfied for a plan to be feasible: HC-01 employee competencies match the assignment requirement; HC-02 employee is on duty (within the weekly schedule, not on time-off, not already assigned); HC-03 vehicle is in active status; HC-04 vehicle type matches the assignment requirement; HC-05 no employee is double-booked; HC-06 no vehicle is double-booked. Five soft constraints encode preferences with configurable weights: SC-01 prefer assigning an employee to their designated vehicle; SC-02 balance workload evenly across employees; SC-03 minimise transitions between consecutive assignments for an employee; SC-04 schedule high-priority assignments first; SC-05 respect stated employee preferences. Norwegian working-time regulation enters as a separate \emph{legal} tier alongside the hard tier: an eleven-hour daily rest minimum, a forty-eight-hour weekly working-hour bound averaged over a reference period, and a six-consecutive-day work limit, all as defaults from \texttt{DEFAULT\_CONSTRAINTS} that the benchmark runner backfills for legacy datasets.

The objective function combines coverage and soft-constraint compliance under the lex-tier model. The portfolio compares solutions lexicographically: hard, then legal, then business-critical, then preference. The flattened weighted total reported alongside the lex breakdown is kept for human readability but does not drive engine selection. Coverage (the count of assignments actually placed) is rewarded in the medium tier so that an engine that places more assignments at the cost of more soft-constraint violations is not automatically penalised. Workload balance enters the soft tier as a sum-of-squares term across employees, so that a single overloaded driver costs more than several drivers each lightly above average. Per-company configurability of the soft-constraint weights means that the same engines produce different plans for different companies without code changes.

Three known limitations of the algorithm are forwarded to the limitations chapter. The OR-Tools CP-SAT engine does not complete the large instance (five hundred assignments) within the sixty-second time budget; the heuristic-versus-exact crossover point under the current formulation is reported below as a real result and forwarded as part of L9. Pathological soft-constraint weight combinations (weights summing to zero, one weight dominating the others) produce unstable rankings or single-criterion plans; the user-experience challenge of explaining what each weight actually means to a coordinator who does not work in optimisation vocabulary is acknowledged and forwarded to L9. The lex contract's strict ordering by hard violations can mask cases where a metaheuristic is better on the legal, business-critical, and preference tiers but loses on the hard tier; this is reported below as a finding about the scoring contract rather than as a flaw in any single engine.

The benchmark runs the four engines (greedy, OR-Tools, Timefold, auto) against three synthetic datasets: small (twenty assignments, five employees, three vehicles), medium (one hundred assignments, twenty employees, ten vehicles), and large (five hundred assignments, fifty employees, twenty-five vehicles). The reference run on 2026-05-02 produces the following results. On small, all three engines schedule twenty-five per cent of assignments, the greedy and auto totals come in at -15{,}151.20, and OR-Tools and Timefold tie at -15{,}951.20. On medium, the scheduled-assignment percentage rises from forty-three per cent (greedy and auto) to fifty-one per cent (OR-Tools) and fifty per cent (Timefold), with totals of -48{,}493.20 (greedy and auto), -58{,}029.00 (OR-Tools), and -57{,}471.95 (Timefold). On large, OR-Tools times out at sixty seconds without producing a result; greedy and auto schedule 35.6 per cent in nineteen milliseconds at -328{,}975.01; Timefold schedules 39.4 per cent in 45.3 seconds at -306{,}857.47. Auto picks greedy on every dataset because greedy has the smallest hard-tier penalty, even though Timefold beats greedy on the legal, business-critical, and preference tiers on the large instance. The headline engineering result is the Timefold engine moving from being the worst-performing engine pre-rework (an average canonical penalty around -446{,}765 across datasets in the v0 \texttt{HardSoftScore} configuration) to being competitive with the heuristic baseline post-rework (around -126{,}760 average), a reduction of approximately seventy-two per cent traceable to the constraint-provider rework on 2026-04-23. OR-Tools is the strongest individual engine on the instances it can finish (small and medium) but cannot complete the five-hundred-assignment large instance within the sixty-second budget, and that is the operational scaling boundary the thesis reports. Wall-clock figures are single-sample and should be read as order-of-magnitude rather than as production performance numbers; the boundaries are forwarded as L5 and L9 in §5.4.


\section{DSR Artifacts Mapping}
\label{sec:dsr-artifacts}

The DSRM Applied bullets in §3.2 named the activities at the level of the methodology. This section makes them concrete by naming each project artefact and the DSR category it sits in. \Cref{tab:dsr-mapping} maps the artefacts produced by the project to the four DSR artefact categories: Construct, Model, Method, and Instantiation. Constructs are the conceptual vocabulary the project introduces or formalises. Models are the abstractions over the problem domain. Methods are the procedures that operate on models. The Instantiation is the runnable artefact in which constructs, models, and methods are realised together.

\begin{table}[h]
\centering
\caption{Project artefacts mapped to DSR categories.}
\label{tab:dsr-mapping}
\small
\begin{tabular}{p{0.22\textwidth}p{0.22\textwidth}p{0.27\textwidth}p{0.22\textwidth}}
\toprule
\textbf{Construct} & \textbf{Model} & \textbf{Method} & \textbf{Instantiation} \\
\midrule
Three locked anchor concepts (Efficiency, Trust/control, Adaptability). Deviation taxonomy (eight categories with severity and status). Soft-constraint weight schema. & Data model with entity relations and the multi-tenant boundary on \texttt{companyId}. Constraint formulation for the assignment problem (six hard constraints, five soft constraints, separate legal tier). & Multi-engine selection process (lexicographic portfolio across hard, legal, business-critical, preference). Benchmarking framework on synthetic small, medium, and large datasets. HITL override flow (inspect, modify, accept, or reject). Time/quality control (fast / less accurate versus slower / more accurate). & RP as deployed application: Next.js + tRPC + PostgreSQL platform with three optimisation engines (greedy, OR-Tools CP-SAT, Timefold) running as subprocesses, drag-and-drop Gantt-style planning timeline, deviation viewer, dashboard, and per-company soft-constraint weight configuration. \\
\bottomrule
\end{tabular}
\end{table}

The mapping is the concrete instantiation of the DSRM Applied bullets in §3.2 and is referenced from the discussion in Chapter 5 wherever a specific artefact is being argued about.


\section{Process Documentation}
\label{sec:process-documentation}

\subsection{Sprint Summary}
\label{sec:sprint-summary}

The project ran from project kick-off in mid-February 2026 through thesis submission in May 2026, organised as eight named iterations rather than as fixed-duration sprints. The capability timeline runs from the engine framework on 2026-02-16 through the greedy baseline on 2026-02-17, the benchmark harness with the three synthetic datasets on 2026-02-17/18, the OR-Tools CP-SAT engine on 2026-02-19, the Timefold engine on 2026-02-24, the shared constraint configuration on 2026-02-26, work-hour bounds and overlap checks on 2026-04-09, the portfolio engine on 2026-04-14, lexicographic scoring with Norwegian Arbeidsmiljøloven constraints on 2026-04-14, the Timefold engine-correctness rework on 2026-04-23, and the benchmark-runner measurability fix on 2026-05-02. The seven coordinator interviews were conducted across the first half of March 2026.

\subsection{Key Decisions}
\label{sec:key-decisions}

A small number of decisions shape the artefact and the thesis around it, and each is referenced by the §3.5 iteration narrative and by the discussion in Chapter 5. The decision to enforce multi-tenancy at the data layer rather than at the application layer was made at the start of §3.5.1 (Schema and Timeline Scaffolding) and is the architectural reason the per-company configurability in §3.5.8 works as it does. The decision to add OR-Tools CP-SAT as a complete solver alongside the greedy baseline was made in §3.5.3 after the greedy heuristic's locally-good-globally-suboptimal pattern surfaced on instances above thirty assignments. The decision to add Timefold as a third engine was made in §3.5.4 after the OR-Tools scaling boundary at the sixty-second budget surfaced on the large instance. The decision to make override low-friction (drag-and-drop, save-with-warning rather than save-blocked) was made in §3.5.5 because a blocking validator would have told the coordinator that the system holds final authority, which is the configuration Bainbridge warns against and the one Trust/control forbids. The decision to expose the time-quality tradeoff to the coordinator rather than handle it as an internal default was made in §3.5.7 because different planning situations carry different urgency-quality requirements. The decision to make the soft-constraint weights per-company configurable rather than to hard-code them was made in §3.5.8 to deliver Adaptability without code changes per company.

\subsection{Time Tracking}
\label{sec:time-tracking}

Time was tracked by author and task category across the project. Embret carried the system-development workstream: the data model, the planning timeline, the deviation viewer, the dashboard, the three optimisation engines, the benchmark harness, the multi-tenant infrastructure, and the user interface. Mikael carried the user-research workstream: the seven interviews, the thematic analysis, the requirements register, the fit/gap analysis, and the thesis writing. Both authors contributed to discussion, scope decisions, and the eight named iterations. Detailed time-tracking summaries by hours per author per task category are recorded in the project's working notes.
