\chapter{Theory}
\label{chap:theory}

This chapter establishes four theoretical foundations on which the rest of the thesis stands. Resource scheduling under a utilization lens (\Cref{sec:scheduling}) supplies the problem vocabulary and preloads efficiency; a five-layer human-in-the-loop theory (\Cref{sec:hitl}) supplies the configuration vocabulary and preloads trust/control; Transport Management Systems (\Cref{sec:tms}) position the category Ressursplanlegger sits inside and preload adaptability; and Design Science Research (\Cref{sec:dsr}) supplies the research paradigm under which an artefact-plus-knowledge contribution is built and investigated. Section length is asymmetric on purpose: \Cref{sec:hitl} carries the most argumentative load and is the longest, while Vehicle Routing theory inside \Cref{sec:scheduling} and TMS-as-category in \Cref{sec:tms} are referenced narrowly, the former to delimit, the latter to locate the planning gap the artefact fills. Because this is a bachelor's thesis, none of the four bodies is treated exhaustively: the chapter draws from each only as far as is needed to ground the choices made in RP and the claims made about it.

\section{Resource Scheduling}
\label{sec:scheduling}

Many everyday jobs come down to scheduling. A dentist fits one-to-one appointments with their patients into the hours of the working day. A waiter routes the next free server to the next ready table. A parent juggles pickup times across two children's after-school activities. Each is a small version of the same problem: matching time-limited resources to tasks at the right time, under rules about who can do what. \textcite{pinedo2016scheduling} defines scheduling more formally as allocating resources to tasks over time, subject to given constraints, with the goal of optimising one or more objectives. The version that runs at a Norwegian transport company every morning sits inside this definition. A traffic coordinator matches drivers and vehicles to the day's transport assignments, where each assignment has a pickup time, a destination, and rules about who and what can do it. The coordinator's job is to decide who drives which truck, where, and when, so that every assignment is covered by someone qualified and available.

The version of the problem RP solves is harder than the simplest single-resource case, because each transport assignment needs both a driver and a vehicle, and the two must stay paired for the whole assignment. In scheduling vocabulary, assignments are tasks with fixed time windows and fixed pickup-and-delivery locations, while drivers and vehicles are resources with their own attributes: a driver has competencies and availability, a vehicle has type and capacity. The planning horizon is dynamic: the coordinator sets it in the dashboard and can plan a single day, a week, or longer rotations as the company's operating rhythm requires. Pairing makes the problem harder than the single-resource case that dominates \parencite{pinedo2016scheduling}, because a candidate plan now has to satisfy compatibility for two resource types at the same moment, not one.

A feasible plan is not the same as a good plan. The personnel-scheduling literature judges plans not only by whether they are feasible but by how well they use the resources available: how much overtime they generate, how much idle time sits between consecutive assignments, and how evenly the workload spreads across staff \parencite{ernst2004staff}. The thesis groups these dimensions under efficiency, the anchor concept for improved resource utilization. Visibility comes first. A coordinator working from a spreadsheet or whiteboard cannot improve resource utilization without seeing it measured, so the artefact's first contribution is putting those utilization measures in front of the coordinator, before any algorithm reorders the plan.

Scheduling constraints are split into two kinds: hard and soft. Hard constraints must hold for a plan to be feasible at all: every driver-vehicle pairing has to be legally and physically capable of the assignment, and no resource may be double-booked. Concrete cases include the driver's licence class matching the vehicle, the driver holding any required certifications such as an ADR certificate for hazardous cargo, and the driver being on duty rather than on declared time-off. Soft constraints encode preferences the plan should honour where possible but may be traded off. Each is weighted in the objective function the solver tries to minimise. Examples include even workload across drivers, individual driver preferences, and continuity with the same customer over time. \textcite{ernst2004staff} state the rule directly: hard constraints must be satisfied while soft constraints may be violated, the latter entering the objective as penalty terms. \textcite{rossi2006constraint} say the same in constraint-programming vocabulary, distinguishing deadlines that bound feasibility from due dates that enter the weighted objective. The weights on soft constraints are configurable per company in RP, so a coordinator's notion of a good plan can be expressed without code changes.

Finding a guaranteed-best plan for a multi-resource assignment problem belongs to the class of NP-hard problems. NP-hard means that no algorithm is known that can solve every instance of the problem in time that grows polynomially with the size of the input, and one is widely believed not to exist. In practical terms, the time it takes to search every possible plan grows so fast as the number of drivers, vehicles, and assignments grows that exhaustive search stops being practical even on the daily fleets the interviewed companies operate, which range up to dozens of drivers and vehicles \parencite{rossi2006constraint, pinedo2016scheduling}. RP responds to this limit by giving up on provable optimality. Instead, the system uses solvers that return good plans within a time budget the coordinator sets, and \Cref{sec:iter-multiengine} describes how three different solvers were combined to do that. A different but adjacent literature, on Vehicle Routing Problems (VRPs), addresses a related question: a VRP minimises total route cost over a sequence of customer visits delivered by a fleet from a depot \parencite{braekers2016vrp}, treating the order of stops as the main thing to decide. RP works on assignments with fixed locations and fixed time windows, so its open question is who performs each assignment, not in what order their stops are visited. \textcite{braekers2016vrp} explicitly exclude problems that combine routing with scheduling decisions from their review, which is the boundary RP sits on the scheduling side of.

The personnel-scheduling literature converges on three families of solvers for multi-resource assignment problems like the one RP solves: constructive heuristics, constraint programming, and metaheuristics. Mathematical programming is a fourth body that, as \textcite{ernst2004staff} note, dominates the rostering literature in volume, but it is rarely used in isolation on messy real-world instances because the constraints and objectives such instances carry are difficult to express in pure mathematical form. The remaining three families each cover a regime the others do not. Heuristics suit messy real-world objectives that resist clean mathematical formulation, constraint programming dominates on highly constrained instances where any feasible plan is acceptable, and metaheuristics search large solution spaces where exact methods do not scale \parencite[pp.~17--18]{ernst2004staff}. No single family dominates across all instances, which is why \Cref{sec:iter-multiengine} instantiates one engine from each. Constructive heuristics, which \textcite{pinedo2016scheduling} calls dispatching rules, build a plan one assignment at a time according to a priority rule (for example, "schedule the most constrained job first"). They are fast, but they can lock onto a locally good plan and miss a much better plan that would be reachable by reordering earlier choices. Constraint programming formalises the problem as variables and constraints and then searches the space of feasible plans systematically. It uses propagation (deducing forced choices from the constraints) and backtracking (undoing a choice that leads to a dead end), and returns near-optimal plans within a time budget at the cost of higher modelling effort \parencite{rossi2006constraint}. Metaheuristics are general-purpose search strategies that improve a current candidate plan by repeatedly modifying it. Two well-known examples illustrate the family. The tabu search of \textcite{glover1986future} keeps a short-term memory of recent moves to avoid cycling back to them. The late-acceptance hill climbing of \textcite{burke2017late} accepts a new candidate if it is no worse than the candidate seen a fixed number of steps ago. Both let the search escape locally optimal plans on large instances. \Cref{sec:iter-multiengine} picks the speed-quality tradeoff up in concrete terms, reporting how each engine behaves on the same benchmark instances. Solver capability alone, however, says nothing about who reads the resulting plan, who can change it, and on what authority. Section §\ref{sec:hitl} turns to that question.

\section{Human-in-the-Loop Automation}
\label{sec:hitl}

A traffic coordinator who reviews and approves every plan the algorithm produces is operating in what the human factors literature calls a human-in-the-loop configuration. The pattern is familiar from other domains. A nurse confirms a medication dose the system flags. A pilot accepts an autopilot suggestion before the autopilot acts. In each case an automated process generates a recommendation, and a human exercises authority over whether it takes effect. \textcite{parasuraman2000automation} formalise this on a ten-level scale of automation for decision and action selection. The scale runs from level 1 (fully manual) to level 10 (fully autonomous), with intermediate steps that progressively transfer authority from human to machine. Level 4 is "the computer suggests one alternative", level 5 is "executes that suggestion if the human approves", and level 6 is "executes automatically unless the human vetoes within a time window". RP sits at level 5: the solver produces one plan, the plan is shown on the timeline, and no driver is dispatched until the coordinator approves it. The artefact deliberately stops short of level 6: the coordinator must act, not merely fail to object.

The reason for placing partial authority with the coordinator, rather than letting the algorithm act on its own, has a structural answer in \textcite{bainbridge1983ironies}. Bainbridge's central observation is that the people who specify automation are not the same as the people who must operate it. Owners and managers ask for the system; the worker on the floor must run it day to day. This split applies to the present project as well: the system was specified by a software firm whose customers are transport-company owners, while a traffic coordinator must run it every shift. Bainbridge argues that automation built under this asymmetry tends to leave operators with an "arbitrary collection of tasks" and inadequate support for them \parencite[p.~775]{bainbridge1983ironies}. Grounding HITL on this asymmetry rather than on convenience makes trust/control a structural requirement of the system, not a usability feature. The coordinator must be able to inspect, modify, accept, or reject any algorithm-generated assignment, because the system was specified by people who do not bear the consequences of its mistakes.

Trust in the algorithm is a separate question from authority over it. A coordinator who has the authority to override the algorithm may still leave the system unused if they do not trust what it produces. \textcite{hoff2015trust} review 127 studies of human-automation trust from 101 articles, and group their findings into a three-layer model of dispositional, situational, and learned trust \parencite[p.~413]{hoff2015trust}. Dispositional trust reflects how much an individual tends to trust automation in general. Situational trust depends on context and workload. Learned trust is built through experience with the specific system: first as initial trust based on reputation, then as dynamic trust that updates with each interaction \parencite[p.~420]{hoff2015trust}. Two coordinators using the same plan generator may therefore hold different levels of trust at the same moment, and the same coordinator's reliance on the algorithm shifts across days. Hoff and Bashir also report that early errors damage trust more than later errors do \parencite[p.~426]{hoff2015trust}, which has a concrete consequence for deployment: a coordinator's first week with RP weighs more in trust formation than the months that follow. The studies the model rests on are mostly from aviation and military settings, so applying it to a small Norwegian coordinator pool is by analogy rather than direct empirical support. By analogy, the model gives the project not a number to predict but a design constraint: calibrated trust (neither over- nor under-reliance) is the target the artefact must enable through the override and explanation interfaces that the next two layers specify.

The coordinator's ability to interpret what the algorithm has done is what keeps trust calibrated rather than blind. \textcite{miller2019explanation} surveys explanation in the social sciences and argues that explanation has three properties most automated systems ignore. Explanations are contrastive: people ask "Why P rather than Q?", not just "Why P?". They are selective: people expect one or two relevant causes rather than a complete causal chain. They are social: people treat explanation as a communicative act, tailored to what they already know \parencite[p.~3]{miller2019explanation}. For an algorithm-assisted plan, the coordinator's question is rarely "Why was this assignment generated?". It is "Why was this driver assigned, rather than the one I would have picked?". Miller argues that explanation is an interface property of an automated decision system, and that "trust is lost when users cannot understand traces of observed behaviour or decisions" \parencite[p.~3]{miller2019explanation}. The implication for RP is that the system must show contrastive reasons for its choices in the coordinator's vocabulary (competence, time, current load), not in raw constraint weights and solver internals. Explanation belongs alongside the override controls themselves, not as a feature added afterwards.

The fourth and deepest layer in HITL theory is a definition of trust itself. \textcite{lee2004trust} define trust as "the attitude that an agent will help achieve an individual's goals in a situation characterized by uncertainty and vulnerability" \parencite[p.~54]{lee2004trust}, and argue that the design goal is appropriate trust, not greater trust. A coordinator who accepts every plan without review (over-reliance, what Lee and See call \textit{misuse}) and a coordinator who reverts to fully manual planning despite a competent solver (under-reliance, what they call \textit{disuse}) are both failures of trust calibration. Lee and See's design principles for trustable automation specify that the system should reveal its past performance, expose its process, and communicate its purpose in terms that connect to the user's goals \parencite[p.~74]{lee2004trust}. In RP, those three dimensions correspond to deviation history, scoring breakdown, and the explicit framing of the system as decision support rather than decision maker. The threat Lee and See identify is concrete. If coordinators do not understand or trust the algorithm, they will not adopt it, and the visibility gap §\ref{sec:scheduling} framed under efficiency will remain unaddressed regardless of solver quality. Chapter 5 returns to this adoption question.

Parasuraman's automation taxonomy, Hoff and Bashir's trust antecedents, Miller's explanation as interface, and Lee and See's definition of appropriate trust together translate into concrete design constraints on RP's HITL interface. The planning timeline at \texttt{/ressursplanlegger} shows the algorithm's suggested plan in the same Gantt view the coordinator uses for manual work, so the suggestion is open to inspection rather than presented as a finished output. After a solver run, a summary panel reports the solution score and counts of hard and soft violations, which exposes process and performance in the sense Lee and See describe. The deviation viewer at \texttt{/avvik} lists each active conflict by type, affected assignment, and affected resource, so a coordinator who disagrees with the algorithm can locate the contested assignment and decide whether to override. Override is intentionally low-friction: dragging an assignment block to a different employee row reassigns it, and conflicting changes save with a warning rather than a block. That choice is deliberate. A blocking validator would tell the coordinator that the system, not the coordinator, holds final authority, which is precisely the configuration Bainbridge warns against and the one Trust/control forbids. Soft constraints are weighted per company so that a coordinator's notion of a good plan can be adjusted without code changes, and the choice of solver engine and time budget is exposed at the point of plan generation so the system's capability limits are visible before the plan is read. These mechanisms are not features added to a finished automation. They are what makes the automation operable at all. \Cref{sec:iter-hitl,sec:iter-deviations,sec:iter-tradeoff} document how each layer was built into the artefact.

\section{Transport Management Systems}
\label{sec:tms}

The first thing to establish about RP's environment is what kind of software the coordinator already works inside or alongside. Transport companies run their day-to-day operations on a category the industry groups under one label: a Transport Management System (TMS). \textcite{heinbach2022datadriven} define a TMS as software that plans and executes the physical movement of goods across supply chains \parencite[p.~811]{heinbach2022datadriven}, a definition the same authors trace back to TMS as freight-forwarder software for assigning customer shipments to fleet resources. \textcite{griffis2007tms} give a compatible definition from a North American shipper survey: a TMS is information technology used to plan, optimise, and execute transportation operations before, during, and after the physical movement of goods \parencite[p.~19]{griffis2007tms}. Across the rest of this section, "TMS" covers four operational areas: order management, route handling, carrier coordination, and freight billing.

Inside that label, the picture is not uniform. What a TMS actually offers depends on the vendor, and whether a transport company uses one at all depends on the company. In the Norwegian market, the two commercial systems named in the interviews are \textbf{Timpex} and \textbf{Opter}. Both cover order management and freight billing, and neither generates assignment plans automatically. Looking beyond order and invoicing tools, \textcite{heinbach2022datadriven} catalogue eleven European platform providers and identify TRUCKAHEAD as a transport workflow platform that combines dispatcher–driver communication with workflow automation \parencite[p.~819]{heinbach2022datadriven}. Adoption itself is uneven. \textcite{griffis2007tms} report that of the firms in their 2007 North American sample without any TMS, half manage transportation entirely manually, while the other half rely on partial or legacy IT \parencite[p.~26]{griffis2007tms}. The sector that RP enters is therefore a patchwork: different vendors cover different ground, and a substantial share of operations run without dedicated planning software at all.

Even where a TMS is in place, one part of the operation is missing from the category as a whole: the upstream planning step that decides which driver and vehicle perform which assignment. \textcite{griffis2007tms} observe that most TMS offerings centre on the individual shipment as the unit of analysis and lack the ability to optimise multi-load shipments, leaving load consolidation as a manual activity \parencite[p.~27]{griffis2007tms}. \textcite{heinbach2022datadriven} reach the same conclusion: TMS products focus on shipment-level business activities, while other digital platforms add capabilities outside that focus \parencite[p.~808]{heinbach2022datadriven}. What existing systems cover, in short, is order management, freight billing, and shipment-level execution. What is missing is the upstream planning step between an order existing and a specific driver and vehicle being assigned to perform it. RP is positioned in exactly that gap. \Cref{chap:findings} substantiates the gap with the empirical comparison, and the per-company configurability that lets RP fit different vendor environments is built in \Cref{sec:iter-weights} and discussed under adaptability in §5.1.3. Section §\ref{sec:dsr} turns to the research paradigm under which an artefact filling the planning gap is built and investigated.

\section{Design Science Research}
\label{sec:dsr}

Design Science Research (DSR) is the research paradigm in which a designed artefact is constructed and investigated as the central contribution of a study, together with the knowledge produced about how that artefact behaves in its context. The point is that the artefact and the knowledge come out of the same project. \textcite{wieringa2014dsm} states the position concisely: design science is the design and investigation of artefacts in context, where the artefact is built to improve something in its problem context \parencite[p.~3]{wieringa2014dsm}. \textcite{hevner2007threecycle} structures the same work as three linked cycles. The relevance cycle imports requirements and acceptance criteria from the application environment. The rigor cycle draws on the existing scientific knowledge base and contributes new knowledge back to it. The design cycle iterates internally between constructing and evaluating the artefact \parencite[pp.~88--90]{hevner2007threecycle}. \textcite{peffers2007dsrm} structure the same paradigm as a procedural model. Their Design Science Research Methodology (DSRM) specifies six activities: problem identification and motivation; objectives for a solution; design and development; demonstration; evaluation; and communication \parencite[p.~46]{peffers2007dsrm}. The two framings are complementary: the cycles describe what must be present, the activities describe what is done. Chapter 3 applies the DSRM activities step by step to this project.

DSR is not the only research paradigm available for an information systems study, and two alternatives in the IS literature warrant comparison. \textcite{orlikowski1991studying} survey the IS literature and group it into three philosophical stances: positivist, interpretive, and critical \parencite[p.~5]{orlikowski1991studying}. The positivist stance assumes an objective social reality investigated with structured instruments, with the goal of testing hypotheses to increase predictive understanding \parencite[p.~5]{orlikowski1991studying}. That stance fits a study that measures population-level effects of a deployed system, but on its own it cannot justify building a new artefact whose primary contribution is the design knowledge embedded in it. The interpretive stance assumes that organisational reality is socially constructed and seeks contextual understanding through participants' meanings rather than generalisable laws \parencite[p.~5]{orlikowski1991studying}. That stance fits a study whose contribution is rich description of coordinator practice, but on its own it cannot support building a runnable artefact whose behaviour can be checked against external criteria. DSR is the third paradigm the present project requires: an artefact-plus-knowledge contribution in which qualitative input informs requirements, and the artefact itself is examined against those requirements.

Inside DSR, the distinction between validation and evaluation shapes what kind of empirical evidence the project can offer. \textcite{wieringa2014dsm} distinguishes validation from evaluation. Validation predicts whether a deployed system would do what stakeholders want; evaluation observes a deployed system in actual use \parencite[p.~31]{wieringa2014dsm}. In validation, the researcher experiments with a model of how stakeholders would use the artefact. In evaluation, stakeholders use the artefact independently of the researcher \parencite[p.~34]{wieringa2014dsm}. The present thesis validates rather than evaluates RP. Behaviour is predicted in two ways. The first is benchmarking on synthetic datasets, which is a single-case mechanism experiment in Wieringa's classification (one artefact, controlled inputs) \parencite[p.~64]{wieringa2014dsm}. The second is requirements traceability against the requirements derived in the relevance cycle. No transport company has operated the system independently. \textcite{hevner2007threecycle} treats this exact ordering as expected practice, with artefacts tested in laboratory and experimental situations before being released into field testing along the relevance cycle \parencite[p.~91]{hevner2007threecycle}, so the absence of production deployment is a structural feature of the project's stage rather than a methodological deficit. \Cref{sec:evaluation-framework} applies this distinction to the synthetic-benchmark protocol, and Section 5.4 returns to it as limitation L5 when bounding what that validation can and cannot claim.
