\chapter{Discussion}
\label{chap:discussion}

% See context/outline.md Ch 5 for paragraph-level plan.

This chapter interprets what \Cref{chap:findings} reported. It organises the primary findings under the three locked anchor concepts (Efficiency, Trust/control, Adaptability), uses the layered theory built in \Cref{chap:theory} to read those findings, and then names the limitations that bound what the thesis can claim. \Cref{sec:primary-findings} carries the core argument under the anchors. \Cref{sec:adoption} draws out adoption implications. \Cref{sec:sustainability} works through sustainability and ethical considerations. \Cref{sec:limitations} catalogues the twelve named limitations (L1 through L12) hierarchically. \Cref{sec:deviations} names deviations from the original plan. \Cref{sec:methodology-reflection} closes with a self-critical reflection on the central methodological weak spot.


\section{Primary Findings under the Anchors}
\label{sec:primary-findings}

The discussion organises primary findings under the three locked anchor concepts: Efficiency, Trust/control, and Adaptability, one sub-section per anchor, with the names used verbatim throughout.

\subsection{Efficiency}
\label{sec:efficiency}

The primary surprising result of \Cref{sec:interview-findings} is the resource-utilization visibility gap. Across the seven interviewed companies, overtime is handled reactively, idle time between assignments is invisible in current tools, and load balancing across drivers depends on coordinator memory rather than on any structured overview. The gap is not, on the face of it, a technological problem. The data needed to see utilization (driver hours, assignment durations, vehicle activity) exists in every company. What is missing is a structured view that makes utilization patterns legible to the people who would act on them.

The asymmetry that makes this gap a finding worth surfacing is the operator-vs-owner asymmetry. Owners and Admmit articulate utilization optimisation as something the company needs. Coordinators do not articulate it on their own. They ask for speed, override authority, and a system that does not get in their way. This is the configuration \textcite{bainbridge1983ironies} anticipates: demand for automation is articulated by those who design and own the system, while the residual operator role is left to whoever must run it day to day \parencite[p.~775]{bainbridge1983ironies}. Bainbridge's argument was theoretical when introduced in §2.2; the seven Norwegian transport interviews extend it empirically. The asymmetry is not an artefact of the interview sample: it appears across companies of different size, different current system, and different geographic location. The thesis does not claim the asymmetry is universal in Norwegian transport, but reports it as the consistent configuration in the consultation pool.

Three concrete utilization dimensions sit under \textbf{Efficiency} as the anchor names: overtime, idle time between consecutive assignments, and uneven load between drivers. RP makes each of them visible before any algorithm reorders them. The planning timeline shows assignments laid out across drivers and vehicles in a Gantt view; gaps between assignments are visible as gaps; long days with multiple consecutive assignments are visible as long bars; uneven load between drivers is visible as visual asymmetry across rows. The deviation viewer surfaces overtime violations as structured conflict data tied to specific assignments. The dashboard at \texttt{/oversikt} aggregates utilisation metrics across drivers and vehicles. Visibility itself is the precondition for optimization: a coordinator working from a spreadsheet or whiteboard cannot improve what they cannot see, and an algorithm cannot optimise what is not formalised.

The multi-engine benchmark in §4.5 is the validation instrument designed for Efficiency. It tests how solver approaches compare under realistic constraint combinations on identical synthetic problem instances; it does not test whether the visibility gap is real (the interviews surfaced this) or whether HITL is necessary (Admmit fixed this from project start). This how-not-whether framing is the methodologically independent test the thesis can defend. The headline result, traceable to commit \texttt{43c6124} on 2026-04-23, is the Timefold engine moving from being the worst-performing engine in its initial \texttt{HardSoftScore} configuration to being competitive with the heuristic baseline post-rework, an approximately seventy-two per cent reduction in canonical penalty across the three datasets. OR-Tools CP-SAT is the strongest individual engine on the instances it can finish (small and medium) but cannot complete the five-hundred-assignment large instance within the sixty-second time budget. That is the operational scaling boundary the thesis reports under Efficiency.

\subsection{Trust/control}
\label{sec:trust-control}

The argument that the coordinator should keep authority over every algorithm-generated assignment rests on Bainbridge's owner-vs-operator framing introduced in §2.2 and §5.1.1. Operator authority over override is not a usability feature added on top of an automated planner; it is the resolution to the asymmetry that motivated automation in the first place. \textcite{parasuraman2000automation} situate this configuration on their ten-level scale of automation for decision and action selection, where level 5 is "executes that suggestion if the human approves" and level 6 is "executes automatically unless the human vetoes within a time window". RP sits at level 5: the solver produces one plan, the plan is shown on the timeline, and no driver is dispatched until the coordinator approves it. Stopping deliberately at level 5 rather than going to level 6 makes override the constitutive interaction rather than the exception, which is what Trust/control requires.

Whether a coordinator will trust the algorithm enough to use it is a separate question from whether they should keep authority. \textcite{hoff2015trust} review 127 studies and group their findings into a three-layer model of dispositional, situational, and learned trust \parencite[p.~413]{hoff2015trust}. Dispositional trust reflects how much an individual tends to trust automation in general. Situational trust depends on context and workload. Learned trust is built through experience with the specific system, first as initial trust based on reputation and then as dynamic trust that updates with each interaction \parencite[p.~420]{hoff2015trust}. \textcite{lee2004trust} sit beneath this three-layer model with the older claim that the design goal is appropriate trust, not greater trust: a coordinator who accepts every plan without review and a coordinator who reverts to fully manual planning despite a competent solver are both failures of trust calibration \parencite[p.~74]{lee2004trust}. Hoff and Bashir refines, rather than replaces, Lee and See: the three-layer model gives the developer a more granular set of antecedents to design against. Hoff and Bashir's finding that early errors damage trust more than later errors do \parencite[p.~426]{hoff2015trust} has a concrete consequence for RP: a coordinator's first week with the system weighs more in trust formation than the months that follow, and the deviation viewer's job in that first week is to make the algorithm's reasoning legible enough that an early misalignment can be inspected and corrected rather than read as a system fault.

Calibration depends on the coordinator being able to interpret what the algorithm has done, and on the system explaining itself in the coordinator's vocabulary rather than in solver internals. \textcite{miller2019explanation} argues that explanations are contrastive ("Why P rather than Q?"), selective (one or two relevant causes rather than a complete causal chain), and social (tailored to what the recipient already knows) \parencite[p.~3]{miller2019explanation}. RP shows contrastive reasons through the deviation viewer (each conflict flags type, affected assignment, and affected resource), through the post-solve summary panel (solution score and counts of hard and soft violations), and through the time-quality control surface introduced in §3.5.7 that labels the toggle as "fast suggestion" versus "best-effort plan" rather than as bare time labels. \textcite{amershi2019guidelines} complement Miller's social-science framing with eighteen practical guidelines for human-AI interaction, organised across initial usage, interaction over time, and inevitable system error. Three of those guidelines map directly onto the artefact: making clear what the system can do (the planning timeline shows the solver's suggested plan in the same view used for manual work), making clear how well the system can do what it does (the score panel and the deviation viewer expose process and performance), and supporting efficient correction (drag-and-drop override that saves with a warning rather than a block).

The fourth element under Trust/control is tacit knowledge as the coordinator's irreducible role. The interviews in §4.1 named tacit-knowledge dependency as the second most frequent pain point: every interviewed coordinator holds operational knowledge that is not formally captured by their current system, including who has the right competence on which routes, who is approaching overtime limits, and which vehicle suits which job. RP partially externalises this through the structured data the schema captures (competencies, certifications with expiry dates, work schedules, vehicle types and capacity) and through the deviation taxonomy that turns memorised constraints into structured conflict data on the timeline. What it cannot externalise is the coordinator's situated judgement: which route is harder in winter, which driver handles a difficult customer well, when to deviate from the algorithm because the day is unusual. The four operationalisations of Trust/control (inspect, modify, accept, or reject) are how that irreducible judgement enters the system. Vague control language ("human oversight", "operator supervision") would not do the work the four concrete actions do.

\subsection{Adaptability}
\label{sec:adaptability}

The seven interviewed companies operate under materially different rules. Bergen Bulk Transport has eight to twenty drivers and plans entirely manually. Norlog Tana runs around five routes per day across food-grade temperature-controlled transport with scanning competence requirements. Ottem operates forty-five vehicles across three departments with about a minute per assignment decision and unpredictable order cancellations. Nordic Crane runs crane and transport divisions with a planning horizon stretching from five hours to two months. Different fleet sizes, different operational rules, different willingness to pay. \textcite{venkatesh2003utaut} formalise this kind of variation as differences in adoption thresholds across organisational contexts; their UTAUT (Unified Theory of Acceptance and Use of Technology) model identifies four constructs (performance expectancy, effort expectancy, social influence, facilitating conditions) that shape technology adoption and use. The thesis does not test UTAUT empirically against the seven companies, but the model frames why a single artefact must accommodate genuinely different operational configurations to function meaningfully across them.

The frequency-ranked assignment-criteria list in §4.1 is the same across the seven companies (availability and position, experience, driving and rest-time status, competence and licence, route familiarity, equipment match, overtime status), but the relative weight each company places on each criterion differs. Some companies prioritise workload balance highly. Others prioritise vehicle--driver continuity or route familiarity. The variation is material rather than cosmetic: a plan that is good for one company can be operationally wrong for another, even when both face the same set of orders. The fit/gap analysis in §4.3 reinforces the point at the system level. The fourteen gap items are universal across the seven companies (planning view, automatic plan generation, conflict detection, capacity overview), but the five fit items (order management, billing, driver notification, basic data storage, status tracking) are met differently by Timpex, by Opter, by custom systems, or not at all, depending on the company.

Configurable soft-constraint weights are the technical mechanism that delivers Adaptability inside RP. Each of the five soft constraints (workload balance, vehicle preference, transitions, priority, employee preferences) has a per-company weight that flows into the CP-SAT objective function, so that the same engine produces different plans for different companies without code changes. The mechanism is the kind of variability point that \textcite{mietzner2009variability} describe in their multi-tenant SaaS variability framework: a single deployed instance carries a controlled set of configuration points that distinguish customers without forking the codebase. RP applies this to the soft-constraint layer of the planning problem: the weights are the per-tenant variability points; the engines, the data model, and the HITL interface are shared across tenants.

What is honest to acknowledge is that configurability for soft weights is a partial realisation of Adaptability. Hard constraints (HC-01 through HC-06) are hard-coded, the status workflow definition (planned, approved, in-progress, completed, cancelled) is hard-coded, and per-company assignment-rule taxonomies are hard-coded. A company whose hard constraints differ materially from the six implemented (for example, a company with regulatory rules outside Arbeidsmiljøloven, or a domain where double-booking is sometimes legitimate) cannot be served by configuration alone. Broader configurability of those layers is named as a gap in §5.4 and forwarded as future work in §6.3. Adaptability is also distinct from skalerbarhet, which concerns volume rather than meaningful operation across material rule differences. The thesis does not claim that RP scales to thousands of drivers; it claims that RP can serve materially different operational rulebooks within its current scope through configuration rather than code.


\section{Adoption and Deployment Implications}
\label{sec:adoption}

What the artefact must demonstrate before companies adopt it is a cost-benefit threshold rather than a feature checklist. \textcite{venkatesh2003utaut} model adoption as the result of performance expectancy (will the system improve my work?), effort expectancy (will it be easy to use?), social influence (do people I respect use it?), and facilitating conditions (do the resources to use it exist?). The interviews in §4.1 mapped onto this directly: the sceptical interviewees (Harlem Solutions, Ottem, Bergen Bulk Transport) ranked cost-benefit and ease of use highest; the positive interviewees (the three Norlog branches) ranked performance and integration with existing systems highest. Bergen Bulk Transport's threshold for considering a TMS at all is fleet size: "I'll think about a TMS once we hit fifteen to twenty trucks." That is a performance-expectancy threshold expressed in fleet terms.

The TMS-as-category framing developed in §2.3 returns here as a deployment question. Existing systems (Timpex, Opter, internal tools) cover order management, invoicing, and driver notification. RP covers the planning step. They are complementary, not competing: a transport company that adopts RP keeps its existing TMS for invoicing. The most frequently named missing feature across interviews is integration with the existing billing system. Norlog Lakselv, Norlog Mo i Rana, and Harlem Solutions all flagged invoicing integration as an adoption-shaping requirement. The thesis acknowledges this gap honestly: billing integration is documented in §4.2 as FK-39, marked as Won't for this version, and recommended as future work in §6.3 \parencite{griffis2007tms, heinbach2022datadriven}.

What it would take to move from validation (this thesis) to evaluation (a deployed system used independently of the researchers) is the question §6.3 takes up. The minimum viable deployment is a production pilot with a real Norwegian transport company that bridges the gap between the synthetic-benchmark validation in §4.5 and Wieringa's evaluation criterion (\Cref{sec:dsr}). The pilot would address L5 (synthetic benchmarks, not production data) and L6 (no real-world deployment) at the same time, and would let the artefact's Efficiency claims be tested against operational utilization data rather than against synthetic instances. A second adoption barrier the interviews surfaced is performance: Timpex is described as extremely slow under concurrent use. RP's serverless deployment on a modern Next.js stack makes performance a differentiator rather than a parity feature, and the IFK-04 page-load target (under three seconds) and IFK-05 concurrent-user target (five to fifty coordinators per company) are deliberately chosen to clear the threshold the interviewees implicitly set.


\section{Sustainability and Ethical Considerations}
\label{sec:sustainability}

The Sustainability Awareness Framework (SusAF) developed by \textcite{becker2015karlskrona} and operationalised by \textcite{duboc2020requirements} structures sustainability analysis across five dimensions: technical, economic, social, individual, and environmental. The framework asks designers to consider both immediate effects and longer-term, indirect effects across each dimension. For an algorithm-assisted planning platform that intermediates the relationship between owners, coordinators, drivers, and customers, the framework is appropriate because the artefact's effects are not all in the same dimension and not all on the same time horizon.

The technical dimension covers the artefact's longevity: the multi-tenant architecture and the pluggable solver registry are the elements that support technical sustainability. New solvers can be added without touching the application; new companies can be added without touching the codebase. The economic dimension covers the artefact's value to the operating company: visible utilisation metrics support both internal cost control (overtime, vehicle utilisation) and external pricing decisions, and the per-company configurability supports a multi-tenant deployment model that distributes development cost across customers. The social dimension covers the artefact's effects on the people working with and around it: the HITL configuration deliberately keeps the coordinator's judgement in the loop rather than replacing it, but the system's optimisation of workload balance can also be read as tightening the workload baseline a driver is expected to meet. The individual dimension covers the coordinator's professional identity: a planner whose tacit knowledge has been the source of their value can experience formalisation as a deskilling threat, even when the formalisation was meant to remove key-person risk. The environmental dimension covers the artefact's effects on physical resource use: better utilisation reduces idle time and unnecessary trips, which translates to lower fuel use and emissions, but the magnitude of that effect on the seven interviewed companies is not measured by the thesis and cannot be claimed without further empirical work.

Four ethical dilemmas tie sustainability to specific design questions. Fairness, in the sense \textcite{mittelstadt2016algorithms} survey for algorithmic decision-making, becomes a question about which drivers the algorithm prefers and why: a workload-balance term that smooths total hours can systematically advantage or disadvantage individual drivers depending on the priority weighting and the assignment distribution. Accountability, as \textcite{martin2019accountability} argue, is the question of who is answerable when an algorithm-generated assignment goes wrong, and the answer in RP ties directly to \textbf{Trust/control}: the coordinator who inspected, modified, accepted, or rejected the assignment is the answerable party, not the algorithm or the company that built it. Privacy is the question of how employee data (competencies, work schedules, time-off, sick-leave, performance) is handled inside the system; RP keeps this data scoped to the employing company through multi-tenant isolation, but a richer feature set (driver-facing app, GPS tracking, real-time location) would push privacy into a different regime. Working conditions tie to \textbf{Efficiency}: \textcite{lee2018understanding} document how algorithmic management can compress task slack and tighten worker discretion in ways that erode the qualitative experience of work, and the question for RP is whether visibility into utilisation drives systemic optimisation that benefits the company at the cost of the drivers being optimised over.

Mapping these effects onto the United Nations 2030 Agenda \parencite{un2015agenda2030} is illustrative rather than exhaustive. SDG 8 (Decent work and economic growth) is the dimension where Efficiency improvements (lower overtime, more even workload) and the working-conditions dilemma cut in opposite directions. SDG 9 (Industry, innovation and infrastructure) is where the multi-tenant SaaS deployment model and the open-source solver components fit. SDG 13 (Climate action) is where lower idle time and fewer unnecessary trips are relevant, with the caveat that the magnitude is not measured.

Two more notes. The operator-vs-owner asymmetry that motivates the artefact is itself an ethics question: whose efficiency, whose autonomy? The thesis comes down on the side that lets coordinators keep authority over individual assignments while making utilisation visible to owners, but a different design that pushed authority upward (an automatically-deployed plan that coordinators can only veto in time-critical windows) would be a different ethical commitment. And the sustainability analysis here is analytical, not empirical: it identifies effects the artefact could plausibly have, but does not measure them on the seven interviewed companies. That bound is named as part of L5 and L6 in §5.4.


\section{Limitations}
\label{sec:limitations}

The L1 through L12 names below are locked verbatim from the limitations register; synonyms (e.g., "smaller sample" for L1) drift the spine and must be flagged by reviewers. The mapping from limitations to sub-questions is:

\begin{itemize}
\item \textbf{SQ1} → §5.1.1 (Efficiency) — bounded by L1, L2, L3, L4
\item \textbf{SQ2} → §5.1.2 (Trust/control) — bounded by L8, L10
\item \textbf{SQ3} → §5.1.3 (Adaptability) and cross-anchor — bounded by L5, L6, L7, L9, L11, L12
\end{itemize}

\subsection{Empirical Foundation}
\label{sec:limits-empirical}

\paragraph{L1 — Sample size (7 interviews).} Seven semi-structured calls of approximately fifteen minutes each is a small empirical foundation. The sample was chosen for thematic coverage rather than for statistical generalisation, and the analysis in §3.4 explicitly treats the themes as bounded to the seven interviewed companies rather than as a basis for population-level claims. What L1 prevents the thesis from claiming is anything quantitative about Norwegian transport at scale: it cannot say that the visibility gap holds in eighty per cent of Norwegian transport companies, only that it surfaced consistently across the seven calls in the consultation pool. The brief duration compounds this: the calls were long enough to surface the visibility gap and the operator-vs-owner asymmetry but not long enough to probe edge cases or to test whether a coordinator's articulated need shifts under sustained reflection.

\paragraph{L2 — Self-selection bias in interview sample (Admmit customers only).} The seven coordinators were recruited through Admmit's customer roster rather than through a public call. Admmit's customers are companies that have already chosen to engage with a software vendor in the planning space, which is a self-selected slice of Norwegian transport. What L2 prevents the thesis from claiming is anything about transport companies that have explicitly chosen not to engage with planning software, including a substantial share of small operators that, per Bergen Bulk Transport's own articulation, will not consider a TMS until the fleet exceeds fifteen to twenty trucks. The operator-vs-owner asymmetry in particular may be sharper inside Admmit's customer base than across the broader population, because the asymmetry is part of why owners brought Admmit in.

\paragraph{L3 — Author affiliation with Admmit.} One author has a prior affiliation with Admmit that predates the project. \textcite{malterud2001lancet} treats preconceptions as separate from bias unless the researcher fails to disclose them \parencite[p.~484]{malterud2001lancet}; the affiliation is declared in §3.7 and is named here as a limitation that bounds reflexivity. What L3 prevents the thesis from claiming is full independence from Admmit's existing positions: the prior expectation that a visibility gap might exist, and the prior commitment to HITL, both predate the interviews. The author affiliation does not invalidate the empirical findings, but it does mean that the asymmetry between Admmit's articulation and the coordinators' articulation should be read as a triangulation point, not as an unmediated discovery.

\paragraph{L4 — Interview-guide coverage gap.} The five guide topics (current tools and workflow, assignment criteria, sick-leave handling, attitudes to automation, adoption criteria) covered the questions the project most needed answered, but they did not exhaustively probe what would belong in a richer Adaptability evaluation. Specifically, the guide did not ask coordinators to enumerate the hard constraints they consider non-negotiable beyond licence and competence, did not probe the status workflow definitions across companies, and did not ask about per-company assignment-rule taxonomies. What L4 prevents the thesis from claiming is that the configurability gap named in §5.1.3 (hard constraints, status workflow, assignment rules remain hard-coded) is small. A richer interview guide might have surfaced rule differences large enough to reposition the gap as central rather than peripheral.

\subsection{Validation and Artefact}
\label{sec:limits-validation}

\paragraph{L5 — Synthetic benchmarks, not production data.} The benchmark in §4.5 runs against three synthetic datasets generated to span the fleet ranges named in §3.3. Synthetic data has known properties (controlled overlap density, controlled constraint intensity) that production data does not have. What L5 prevents the thesis from claiming is anything about how RP would behave on real Norwegian-transport operational data. The synthetic generator may understate the constraint pressure of a real day and overstate it on others; the patterns of workload balance and route familiarity that real coordinators would recognise are not present. The benchmark is validation evidence in Wieringa's sense (\Cref{sec:dsr}), not evaluation evidence: it predicts how the engines compare under controlled inputs, not how they perform when stakeholders use the artefact in production.

\paragraph{L6 — No real-world deployment.} No transport company has operated RP independently of the researchers. The artefact is implemented and developer-verified; thirty-six of the thirty-seven in-scope functional requirements pass the implementation-and-developer-verification check in the requirements traceability matrix in §3.6. What L6 prevents the thesis from claiming is that the system delivers measurable Efficiency gains under operational time pressure, that the override flow holds up under coordinator workload, or that adoption barriers identified in §5.2 are correctly weighted. \textcite{hevner2007threecycle} treats the lab-before-field ordering as expected DSR practice \parencite[p.~91]{hevner2007threecycle}, so L6 is a bound on the project's stage rather than a methodological deficit; it is, however, the limitation most central to the thesis's external validity.

\paragraph{L7 — No empirical comparison against existing TMS.} The fit/gap analysis in §4.3 compares RP's coverage against what existing systems cover, but it does not benchmark RP against Timpex, Opter, or the internal and custom tools the seven interviewed companies use. What L7 prevents the thesis from claiming is that RP is better than existing alternatives in the same market segment. The differentiator the thesis can defend is structural (RP fills the planning gap that existing systems do not address), not comparative (RP outperforms alternatives on a shared task). Closing this gap would require operational deployment alongside an existing system, which L6 also bounds.

\paragraph{L8 — No user testing with coordinators.} The HITL interface (the planning timeline, the deviation viewer, the time/quality control, the drag-and-drop override) is implemented and developer-verified, but it has not been user-tested with traffic coordinators independently of the development team. What L8 prevents the thesis from claiming is that the override flow is comprehensible, that the time/quality labels work for coordinators who do not think in solver-runtime budgets, or that the deviation taxonomy surfaces conflicts at a useful threshold. The Trust/control argument in §5.1.2 is design argument plus theory; user testing would convert it to design argument plus theory plus empirical evidence. L8 is the limitation that most directly bounds SQ2.

\paragraph{L9 — Algorithm evaluation against own benchmarks only.} The benchmark in §4.5 reports results against three synthetic datasets that the project itself defined. There is no independent benchmark suite that RP's engines run against, and no public dataset standard for the assignment problem RP solves. What L9 prevents the thesis from claiming is that the engine performance numbers transfer to other formulations of related problems. The Timefold seventy-two per cent canonical-penalty reduction is a real, measurable engineering result against the canonical scorer in this codebase, but cross-codebase comparison would require a shared dataset and a shared scorer that the project does not currently have.

\subsection{Conceptual and Methodological}
\label{sec:limits-conceptual}

\paragraph{L10 — HITL as mandate, not validated level.} HITL was Admmit's requirement from project start. The seven interviews validated the necessity of HITL through coordinator consensus that automated planning must be correctable rather than authoritative, but they did not validate the specific Parasuraman level (level 5: system suggests, human approves) at which RP sits. What L10 prevents the thesis from claiming is that level 5 is the right level for Norwegian transport rather than, say, level 4 (system suggests an alternative, human selects) or level 6 (system executes automatically unless vetoed within a time window). A coordinator who actually used RP under operational time pressure might prefer level 6 for routine assignments and level 5 for exceptional ones; the thesis cannot tell.

\paragraph{L11 — Single-domain (Norwegian transport).} The empirical foundation, the artefact, and the discussion are all anchored in Norwegian transport. What L11 prevents the thesis from claiming is that the visibility gap, the operator-vs-owner asymmetry, the multi-engine architecture, or the configurability mechanism transfer to adjacent domains (other Nordic transport markets, other personnel-scheduling settings such as nurse rostering or airline crew scheduling) without further empirical work. Cross-domain replication would address L11 directly and is named as future work in §6.3.

\paragraph{L12 — Boundary cases described, not quantified.} Several boundary cases in §3.5 (false-positive deviation alerts at week boundaries and mid-shift absences, pathological soft-constraint weight combinations, the lex contract masking Timefold's superiority on the legal and business-critical tiers on the large instance) are described as observations from the iterative development process. They are not quantified across a sample of cases, and the thesis does not report frequency distributions for any of them. What L12 prevents the thesis from claiming is a precise rate at which any boundary case occurs in normal operation. Operational deployment would let these be measured; until then, they remain qualitative observations carried as caveats on the engine results in §4.5.


\section{Deviations from Plan}
\label{sec:deviations}

Three deviations from the original plan are worth naming honestly. First, the seven interviews were originally scheduled to be conducted in a single day. They were instead spread across the first half of March 2026, with one call at a time and reflection between calls so that gaps in earlier conversations could be probed in later ones. The deviation produced a stronger empirical foundation than the single-day plan would have: the visibility gap and the operator-vs-owner asymmetry both surfaced through cross-call comparison rather than from any single conversation. Second, the multi-engine architecture was not in the original scope. The greedy baseline in §3.5.2 was the planned algorithm; OR-Tools CP-SAT was added in §3.5.3 after the greedy heuristic's locally-good-globally-suboptimal pattern surfaced; Timefold was added in §3.5.4 after the OR-Tools scaling boundary surfaced. The deviation produced the methodologically independent how-not-whether benchmark in §4.5 and shifted the thesis's central evaluation from "the system is feasible" to "the system's solver layer is comparable across families". Third, user testing with coordinators was on the original work plan and was not delivered. The decision to forgo user testing was driven by the project's stage (§3.6 frames the work as validation in Wieringa's sense rather than evaluation) and by the limited time available after the eight implementation iterations stabilised. The deviation is named as L8 in §5.4 and as future work in §6.3, and is the most consequential of the three for what the thesis can claim about Trust/control.


\section{Methodology Reflection}
\label{sec:methodology-reflection}

The most load-bearing methodological weak spot in this thesis is the synthetic-benchmark validation that carries Efficiency. The benchmark is the central evaluation instrument for the artefact's primary anchor. It is rigorous within its own terms (reproducible engine versions, shared constraint configuration, lex-tier scoring with Norwegian Arbeidsmiljøloven constraints, single-sample wall-clock figures honestly labelled), and the seventy-two per cent Timefold reduction post-rework is a real engineering result traceable to a specific commit. What the benchmark cannot do is tell the reader how RP behaves on real Norwegian-transport operational data. The synthetic generator was tuned to span the fleet ranges named in §3.3, but synthetic overlap density, synthetic priority distributions, and synthetic time windows are not the same as real ones, and a coordinator using RP on a real day would face constraint pressure the synthetic datasets may understate or overstate. L5 names this bound, L6 names the absence of deployment that would close it, and L7 names the absence of comparison against existing systems that would let the differentiator the thesis claims (planning gap coverage) be tested empirically. If the project had a second year, the most valuable methodological move would be a production pilot with one of the seven interviewed companies, run against a real day's operational data with the coordinator's own override decisions captured. Without that, the thesis's Efficiency claim is structurally a claim about how solvers compare under controlled inputs, not a claim about utilisation gains in deployed Norwegian transport. The thesis is honest about that distinction throughout, but the distinction is the most significant single qualifier on what the work establishes.
